### 2024-11-27 doc.tex
Start of report

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{hyperref}
\usepackage[backend=biber,style=authoryear,autocite=inline]{biblatex}
\usepackage{titlesec}

\titleformat{\section}{\sffamily\Large}{\thesection}{1em}{}
\titleformat{\subsection}{\sffamily\large}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\sffamily\normalsize}{\thesubsubsection}{1em}{}

\addbibresource{ref.bib}

\begin{document}

\section*{AI energy consumption}

\begin{itemize}
    \item \cite{castano2023exploring}
    \item \cite{devries2023growing}
    \item \cite{heguerte2023estimate}
    \item \cite{gowda2024watt}
    \item \cite{luccioni2023counting}
    \item \cite{patterson2021carbon}
    \item \cite{rodriguez2024evaluating}
    \item \cite{tomlinson2024carbon}
    \item \cite{tripp2024measuring}
\end{itemize}

xxxx

\section*{Energy mix}
\begin{itemize}
    \item \cite{krebs2018umweltbilanz}
\end{itemize}

\printbibliography
\end{document}

### 2024-11-27 doc.tex
Device supply chain estimates and other considerations.

@@
 \documentclass[11pt]{article}
+\usepackage[margin=0.7in]{geometry}
 \usepackage[T1]{fontenc}
@@
 \renewcommand{\ttdefault}{fvm} % Use as monospace
+\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
+\newcommand{\assumption}[1]{{\bf #1}}
 %\usepackage{helvet} % Helvetica font for sans-serif
@@
 \usepackage{titlesec}
+\usepackage{enumitem}
 
@@
 \titleformat{\subsubsection}{\sffamily\normalsize}{\thesubsubsection}{1em}{}
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{0.5em}
 
@@
 
-\section*{AI energy consumption}
+\title{UZH IT \& AI Emissions Analysis}
+\date{27.11.2024}
 
+\section*{Introduction}
+
+We are trying to answer two questions:
 \begin{itemize}
-    \item \cite{castano2023exploring}
-    \item \cite{devries2023growing}
-    \item \cite{heguerte2023estimate}
-    \item \cite{gowda2024watt}
-    \item \cite{luccioni2023counting}
-    \item \cite{patterson2021carbon}
-    \item \cite{rodriguez2024evaluating}
-    \item \cite{tomlinson2024carbon}
-    \item \cite{tripp2024measuring}
+    \item Are there emissions savings in moving UZH's IT infrastructure to the cloud?
+    \item What is the impact of increased AI use (and, in particular, of LLMs) on UZH's emissions?
 \end{itemize}
 
-xxxx
+To this end, we need some UZH-specific information:
+\begin{enumerate}
+    \item The size of UZH's student and staff population, and a rough breakdown of
+        what kind of work they do (to roughly estimate their use of IT and AI).
+    \item Information about UZH's current IT infrastructure (number of servers,
+        classroom equipment, etc.), and also, very importantly, the energy utilization of any
+        campus data centers.
+    \item An estimate of the cloud compute capacity required to host UZH's current IT requirements.
+\end{enumerate}
 
-\section*{Energy mix}
-\begin{itemize}
-    \item \cite{krebs2018umweltbilanz}
-\end{itemize}
+In addition, we need more general information about the possible location of cloud data centers and
+LLM deployments, the embodied emissions of different types of IT equipment, and the carbon intensity
+of electricity in different regions.
+
+Below we describe our assumptions and uncertainties regarding these factors, and then describe
+our modeling approach.
+
+\section*{Data sources and unknowns}
+
+\cite{budennyy2022eco2ai}
+\cite{castano2023exploring}
+\cite{devries2023growing}
+\cite{gowda2024watt}
+\cite{heguerte2023estimate}
+\cite{luccioni2023counting}
+\cite{patterson2021carbon}
+\cite{rodriguez2024evaluating}
+\cite{tomlinson2024carbon}
+\cite{tripp2024measuring}
+
+\subsection*{UZH parameters}
+
+We have access to a spreadsheet (\texttt{2024-11-13\_Berechnung\_THG\_IKT\_2024.xls}) with
+counts and emissions factors for various types of IT equipment at UZH. It cites approximately
+10,000 laptops and monitors, 1,000 servers, 2,400 mobile phones,  150 classroom PCs, and a few
+hundred printers, copiers, and projectors.
+
+By contrast, Wikipedia \parencite{wikiuzh} cites approximately 28,000 students and 10,000 staff
+at UZH. Evidently the spreadsheet only lists university assets, not personal devices.
+
+\assumption{In modeling overall university IT emissions as well as potential LLM use,
+we assume that the entire university population has a laptop and a mobile phone.}
+
+We do not yet have any information about the breakdown of different functions or parameters, so
+\assumption{we assume that IT use is uniform across the university}. This assumption needs to be
+revisited.
+
+\subsection*{Embodied emissions}
+The estimation of embodied emissions introduces additional assumptions and uncertainties into
+the model. The secretiveness of major cloud providers makes it especially difficult
+to estimate the embodied emissions of cloud infrastructure. Our initial proposal was to ignore
+embodied emissions altogether, but that leaves a big gap.
+
+Below are some devices and their estimated embodied emissions, from various sources.
+
+\subsubsection*{Laptops}
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|r|l|}
+\hline
+Production \& Disposal & \\
+(kg CO$_2$e) & Source \\ \hline
+121 & Ecoinvent  \\ \hline
+104 & \textcite{teehan2013} (low) \\ \hline
+338 & \textcite{teehan2013} (high) \\ \hline
+244 & \textcite{rarecoil} \\ \hline
+202 & \textcite{unctadder2024} \\ \hline
+\end{tabular}
+\label{tab:embodied_emissions:laptops}
+\end{table}
+
+The Ecoinvent numbers appear to be outliers on the low end.
+
+The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
+information for over 90 popular devices. Mean supply chain emissions are
+244 kg CO$_2$e with a standard deviation of 128 kg. We model laptop emissions using a normal
+distribution with those characteristics.
+
+\subsubsection*{Desktops}
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|r|l|}
+\hline
+Production \& Disposal & \\
+(kg CO$_2$e) & Source \\ \hline
+238 & Ecoinvent \\ \hline
+303 & \textcite{teehan2013} \\ \hline
+403 & \textcite{unctadder2024} \\ \hline
+289 & \textcite{dellpcf} \\ \hline
+277 & \textcite{boavizta} \\ \hline
+\end{tabular}
+\label{tab:embodied_emissions:desktops}
+\end{table}
+
+Most estimates are in the range of 250 - 400 kg CO$_2$e. The most detailed data comes from
+Dell's Product Carbon Footprints (\textcite{dellpcf}), which they publish for each of their products.
+A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
+We model desktop emissions using a normal distribution with those characteristics.
+
+\subsubsection*{Servers}
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|r|l|}
+\hline
+Production \& Disposal & \\
+(kg CO$_2$e) & Source \\ \hline
+383 & \textcite{teehan2013} \\ \hline
+1252 & \textcite{davy2021} (Dell) \\ \hline
+1912 & \textcite{davy2021} (EC2) \\ \hline
+899 & \textcite{boavizta} (medium server)\\ \hline
+\end{tabular}
+\label{tab:embodied_emissions:servers}
+\end{table}
+
+\textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
+The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
+1912 kg with $\sigma = 885$ kg. We model server emissions using the distribution corresponding to Dell
+servers.
+
+\subsection*{Cloud data centers}
+
+\subsection*{Large language models}
+
+\subsection*{Electricity carbon intensity}
+
+\cite{krebs2018umweltbilanz}
+
+\section*{Modeling approach}
+
+\section*{Other considerations}
+
+We typically think of AI use as creating additional energy demands and GHG emissions.
+So far, this model does not include potential energy savings caused by AI, such as improvements
+to building energy management, campus logistics, etc.
+
+\textcite{tomlinson2024carbon} make the argument that tasks such as writing
+and illustrating can be done with fewer GHG emissions by AI than by humans. They compare the
+estimated emissions of an LLM performing the task to those of the average human, calculated
+by multiplying per-capita annual emissions by the time ($\approx 1$h) required to write or draw.
+Using this approach, AI is orders of magnitude cheaper than humans, whether one uses per-capita
+emissions from the US (15 t CO$_2$e/y) or India (1.9 t). Of course, this methodology focuses on
+a specific creative act: the LLM does not reduce the human's overall annual emissions.
 

### 2024-11-28 doc.tex
Add information about phones, other devices, networking, cloud footprint.

@@
 
-\subsection*{UZH parameters}
+\subsection*{UZH population}
 
 We have access to a spreadsheet (\texttt{2024-11-13\_Berechnung\_THG\_IKT\_2024.xls}) with
-counts and emissions factors for various types of IT equipment at UZH. It cites approximately
+counts and emissions factors for various types of IT equipment at UZH. It reports approximately
 10,000 laptops and monitors, 1,000 servers, 2,400 mobile phones,  150 classroom PCs, and a few
@@
 
-By contrast, Wikipedia \parencite{wikiuzh} cites approximately 28,000 students and 10,000 staff
+By contrast, \textcite{wikiuzh} reports approximately 28,000 students and 10,000 staff
 at UZH. Evidently the spreadsheet only lists university assets, not personal devices.
@@
 The estimation of embodied emissions introduces additional assumptions and uncertainties into
-the model. The secretiveness of major cloud providers makes it especially difficult
-to estimate the embodied emissions of cloud infrastructure. Our initial proposal was to ignore
-embodied emissions altogether, but that leaves a big gap.
+the model. Our initial project proposal ignored embodied emissions altogether, but that leaves
+too big a gap.
 
@@
 
-\begin{table}[h]
-\centering
+\begin{center}
 \begin{tabular}{|r|l|}
@@
 \label{tab:embodied_emissions:laptops}
-\end{table}
+\end{center}
 
 The Ecoinvent numbers appear to be outliers on the low end.
-
 The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
 information for over 90 popular devices. Mean supply chain emissions are
-244 kg CO$_2$e with a standard deviation of 128 kg. We model laptop emissions using a normal
-distribution with those characteristics.
+244 kg CO$_2$e with a standard deviation of 128 kg.\
+\assumption{We model laptop supply chain emissions using a normal distribution with those characteristics.}
 
@@
 \label{tab:embodied_emissions:desktops}
-\end{table}
+\end{center}
 
@@
 A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
-We model desktop emissions using a normal distribution with those characteristics.
+\assumption{We model desktop supply chain emissions using a normal distribution with those characteristics.}
 
@@
 \label{tab:embodied_emissions:servers}
-\end{table}
+\end{center}
 
@@
 The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
-1912 kg with $\sigma = 885$ kg. We model server emissions using the distribution corresponding to Dell
-servers.
+1912 kg with $\sigma = 885$ kg.
+\assumption{We model server supply chain emissions using the distribution corresponding to Dell servers.}
+
+\subsubsection*{Smartphones}
+
+\begin{center}
+\begin{tabular}{|r|l|}
+\hline
+Production \& Disposal & \\
+(kg CO$_2$e) & Source \\ \hline
+68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
+54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
+50 & \textcite{unctadder2024} \\ \hline
+50 & \textcite{lovehagen2023} \\ \hline
+\end{tabular}
+\label{tab:embodied_emissions:phones}
+\end{center}
+
+\assumption{We model phones as having mean supply chain emissions of 50 kg CO$_2$e with $\sigma = 10$ kg.}
+
+\subsubsection*{Other devices}
+
+We have found less data for other types devices. For now we model them with averages based
+on a handful of sources.
+
+\begin{description}
+    \item[Computer monitors] 344 kg CO$_2$e (\textcite{teehan2013}, \textcite{dellpcf})
+    \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
+    \item[Printer/copier stations] 1167 kg CO$_2$e (Ecoinvent)
+    \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
+      and have no information about UZH network architecture. The vast majority ($80-95\%$) of
+      lifecycle emissions of network gear stem from usage (\textcite{cisco2024}, \textcite{jacob2023}).
+      For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
+      (A typical data center rack contains 42 1U servers and one switch.)
+\end{description}
 
 \subsection*{Cloud data centers}
+
+The secretiveness of major cloud providers makes it difficult to estimate the embodied emissions of cloud
+infrastructure. The best resource we have been able to find is the Datavizta API (\textcite{boavizta}).
+
+Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as follows:
+\begin{itemize}
+    \item One-for-one replacement of one campus server with one Azure D8S\_v3 instance (2x Intel
+      Xeon, 8 cores, 32 GB RAM, 2TB SSD) (\cite{msftvms}).
+      \item Deployment to Microsoft Azure's ``Switzerland North'' region .
+      \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 kg CO$_2$e for usage,
+      97 kg CO$_2$e for manufacturing.
+\end{itemize}
 
@@
 \section*{Other considerations}
+
+\subsection*{Other IT components}
+
+For now, we do not attempt to model changes in network infrastructure or utilization due to
+move to the cloud or increased LLM use. See \textcite{jacob2023} for an overview of
+network sustainability.
+
+\subsection*{Potential AI benefits}
 

### 2024-11-28 doc.tex
Added operational emissions, LLM emissions, carbon intensity, and many small changes.

@@
 \usepackage{enumitem}
+\usepackage{titling}
 
@@
 \addbibresource{ref.bib}
+\title{UZH IT \& AI Emissions Analysis\vspace{-1em}}
+\author{Massimiliano Poletto}
+\date{\vspace{-1em}27.11.2024}
 
@@
 
-\title{UZH IT \& AI Emissions Analysis}
-\date{27.11.2024}
-
-\section*{Introduction}
+\maketitle
+\section{Introduction}
 
@@
     \item The size of UZH's student and staff population, and a rough breakdown of
-        what kind of work they do (to roughly estimate their use of IT and AI).
+        what kind of work they do (to estimate their use of IT and AI).
     \item Information about UZH's current IT infrastructure (number of servers,
@@
 
-\section*{Data sources and unknowns}
+\section{Data sources and unknowns}
 
-\cite{budennyy2022eco2ai}
-\cite{castano2023exploring}
-\cite{devries2023growing}
-\cite{gowda2024watt}
-\cite{heguerte2023estimate}
-\cite{luccioni2023counting}
-\cite{patterson2021carbon}
-\cite{rodriguez2024evaluating}
-\cite{tomlinson2024carbon}
-\cite{tripp2024measuring}
-
-\subsection*{UZH population}
+\subsection{UZH population}
 
@@
 
-\subsection*{Embodied emissions}
+\subsection{Embodied emissions}
 The estimation of embodied emissions introduces additional assumptions and uncertainties into
@@
 
-\subsubsection*{Laptops}
+\subsubsection{Laptops}
 
@@
 \hline
-Production \& Disposal & \\
-(kg CO$_2$e) & Source \\ \hline
-121 & Ecoinvent  \\ \hline
+\textbf{Production \& Disposal} & \\
+\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
+121 & \textcite{ecoinvent}  \\ \hline
 104 & \textcite{teehan2013} (low) \\ \hline
@@
 information for over 90 popular devices. Mean supply chain emissions are
-244 kg CO$_2$e with a standard deviation of 128 kg.\
-\assumption{We model laptop supply chain emissions using a normal distribution with those characteristics.}
+244 kg CO$_2$e with a standard deviation of 128 kg.
+\assumption{We model laptop supply chain emissions using a truncated normal distribution with those characteristics.}
 
-\subsubsection*{Desktops}
+\subsubsection{Desktops}
 
@@
 \hline
-Production \& Disposal & \\
-(kg CO$_2$e) & Source \\ \hline
-238 & Ecoinvent \\ \hline
+\textbf{Production \& Disposal} & \\
+\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
+238 & \textcite{ecoinvent} \\ \hline
 303 & \textcite{teehan2013} \\ \hline
@@
 A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
-\assumption{We model desktop supply chain emissions using a normal distribution with those characteristics.}
+\assumption{We model desktop supply chain emissions using a truncated normal distribution with those characteristics.}
 
-\subsubsection*{Servers}
+\subsubsection{Servers}
 
@@
 \hline
-Production \& Disposal & \\
-(kg CO$_2$e) & Source \\ \hline
+\textbf{Production \& Disposal} & \\
+\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
 383 & \textcite{teehan2013} \\ \hline
@@
 1912 kg with $\sigma = 885$ kg.
-\assumption{We model server supply chain emissions using the distribution corresponding to Dell servers.}
+\assumption{We model server supply chain emissions using a truncated normal distribution corresponding
+to Dell servers.}
 
-\subsubsection*{Smartphones}
+\subsubsection{Smartphones}
 
@@
 \hline
-Production \& Disposal & \\
-(kg CO$_2$e) & Source \\ \hline
+\textbf{Production \& Disposal} & \\
+\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
 68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
@@
 
-\assumption{We model phones as having mean supply chain emissions of 50 kg CO$_2$e with $\sigma = 10$ kg.}
+\assumption{We model phone supply chain emissions using a truncated normal distribution with mean
+50 kg CO$_2$e and $\sigma = 10$ kg.}
 
-\subsubsection*{Other devices}
+\subsubsection{Other devices}
 
-We have found less data for other types devices. For now we model them with averages based
+We have found less data for other types of devices. For now we model them with averages based
 on a handful of sources.
@@
     \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
-    \item[Printer/copier stations] 1167 kg CO$_2$e (Ecoinvent)
+    \item[Printer/copier stations] 1167 kg CO$_2$e (\textcite{ecoinvent})
     \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
@@
       For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
-      (A typical data center rack contains 42 1U servers and one switch.)
+      (A typical data center rack contains 42 1U servers and one switch or router.)
 \end{description}
 
-\subsection*{Cloud data centers}
+\subsection{Operational emissions}
 
-The secretiveness of major cloud providers makes it difficult to estimate the embodied emissions of cloud
+For on-premise equipment, we would like to model the Swiss electricity mix (see below) and, ideally,
+daily variations in usage and electricity carbon intensity.
+Generic manufacturer estimates of lifetime usage-related emissions are therefore unsuitable.
+Unfortunately, there are few consistent sources of data on device power consumption.
+Based on a smattering of documents from Apple, Cisco, and Dell, we model
+power draw of different devices while under load as truncated normal distributions with the
+following initial parameters:
+
+\begin{center}
+\begin{tabular}{|l|r|r|}
+\hline
+\textbf{Device} & \textbf{Mean (W)} & \textbf{$\sigma$ (W)} \\ \hline
+Laptop & 30 & 5 \\ \hline
+Desktop & 100 & 20 \\ \hline
+Server & 400 & 100 \\ \hline
+Phone & 5 & 2 \\ \hline
+Monitor & 50 & 10 \\ \hline
+Conference display & 120 & 25 \\ \hline
+Printer/copier & 1000 & 200 \\ \hline
+\end{tabular}
+\end{center}
+
+Initially, we assume a distribution of duty cycles and power draw when idle for each device.
+As a next step, we may simulate power draw at different times of day to integrate with
+hourly electricity mix data, but errors in those assumptions may swamp the relatively small
+hourly variations in Swiss electricity carbon intensity.
+
+\subsection{Cloud data centers}
+
+The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
 infrastructure. The best resource we have been able to find is the Datavizta API (\textcite{boavizta}).
@@
       Xeon, 8 cores, 32 GB RAM, 2TB SSD) (\cite{msftvms}).
-      \item Deployment to Microsoft Azure's ``Switzerland North'' region .
+    \item Deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region .
     \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 kg CO$_2$e for usage,
@@
 
-\subsection*{Large language models}
+\subsection{Large language models}
 
-\subsection*{Electricity carbon intensity}
+Energy consumption of LLMs is being studied extensively (\textcite{budennyy2022eco2ai},
+\textcite{castano2023exploring},
+\textcite{devries2023growing},
+\textcite{gowda2024watt},
+\textcite{harding2024watts}, 
+\textcite{heguerte2023estimate},
+\textcite{luccioni2022estimating},
+\textcite{luccioni2023counting},
+\textcite{patterson2021carbon}, 
+\textcite{rodriguez2024evaluating},
+\textcite{tripp2024measuring}).
 
-\cite{krebs2018umweltbilanz}
+The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request,
+or approximately 10x the energy of a traditional search engine query. Indications are that this
+number is decreasing as model hardware and software improve, even though overall energy use is increasing
+due to increased utilization.
 
-\section*{Modeling approach}
+Nevertheless, for the model we assume that queries consume 3 Wh each, and we model per-capita daily
+queries using a truncated normal distribution with mean 15 and $\sigma = 5$.
 
-\section*{Other considerations}
+\subsection{Electricity carbon intensity}
 
-\subsection*{Other IT components}
+The Swiss Federal Office for the Environment published an extensive report (\textcite{krebs2018umweltbilanz})
+on the environmental impact of electricity generation in Switzerland in 2018.
+
+However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
+we use the carbon emissions for electricity {\em consumption} in Switzerland for 2023 at hourly
+granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
+hour block over the course of the year. This gives us twelve data points that describe average
+carbon intensity variation over the course of the day.
+
+\section{Modeling approach}
+
+\section{Other considerations}
+
+\subsection{Other IT components}
 
@@
 
-\subsection*{Potential AI benefits}
+\subsection{Potential AI benefits}
 

### 2024-11-29 doc.tex
Added modeling approach, missing data, etc.

@@
 \author{Massimiliano Poletto}
-\date{\vspace{-1em}27.11.2024}
+\date{\vspace{-1em}29.11.2024}
 
@@
 \begin{enumerate}
-    \item The size of UZH's student and staff population, and a rough breakdown of
+    \item The size of UZH's student and staff population and a rough breakdown of
         what kind of work they do (to estimate their use of IT and AI).
     \item Information about UZH's current IT infrastructure (number of servers,
-        classroom equipment, etc.), and also, very importantly, the energy utilization of any
+        classroom equipment, etc.) and the energy use (electricity bills) of any
         campus data centers.
@@
 
-In addition, we need more general information about the possible location of cloud data centers and
+In addition, we need information about the possible location of cloud data centers and
 LLM deployments, the embodied emissions of different types of IT equipment, and the carbon intensity
@@
 
-Below we describe our assumptions and uncertainties regarding these factors, and then describe
-our modeling approach.
+We describe our modeling approach and our assumptions and uncertainties regarding these factors.
 
@@
 
-We do not yet have any information about the breakdown of different functions or parameters, so
-\assumption{we assume that IT use is uniform across the university}. This assumption needs to be
+We do not yet have any information about the breakdown of different functions / job roles, so
+\assumption{we assume that IT use is uniform across the university}. This assumption should be
 revisited.
@@
 
-Below are some devices and their estimated embodied emissions, from various sources.
+We did some research on the estimated embodied emissions of different types of devices.
 
@@
 
-We have found less data for other types of devices. For now we model them with averages based
+We found less data for other types of devices. For now, we model them with averages based
 on a handful of sources.
@@
 Generic manufacturer estimates of lifetime usage-related emissions are therefore unsuitable.
-Unfortunately, there are few consistent sources of data on device power consumption.
-Based on a smattering of documents from Apple, Cisco, and Dell, we model
+There are surprisingly few sources of data on device power consumption (as opposed to CO$_2$ emissions).
+Based on a smattering of data sheets, primarily from Apple, Cisco, and Dell, we model
 power draw of different devices while under load as truncated normal distributions with the
@@
 \subsection{Cloud data centers}
+\label{sec:cloud_data_centers}
 
@@
 The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request,
-or approximately 10x the energy of a traditional search engine query. Indications are that this
-number is decreasing as model hardware and software improve, even though overall energy use is increasing
-due to increased utilization.
+or approximately 10x the energy of a traditional search engine query. This number appears to be decreasing
+rapidly as model hardware and software improve, even though overall energy use is increasing due to
+increased utilization.
 
@@
 queries using a truncated normal distribution with mean 15 and $\sigma = 5$.
+
+LLMs are distributed and the location that serves any particular request is unknown, so we use the 
+average European electricity mix to compute carbon intensity.
 
@@
 However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
-we use the carbon emissions for electricity {\em consumption} in Switzerland for 2023 at hourly
-granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
+we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
+at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
 hour block over the course of the year. This gives us twelve data points that describe average
@@
 \section{Modeling approach}
+
+At a high level, we model IT emissions primarily bottom-up, based on device counts and
+characteristics, and use a Monte Carlo approach to account for uncertainties.
+
+For on-campus data centers, we would like to use end-to-end data on electricity use (electric bills)
+to evaluate our model.
+
+\subsection{Campus emissions}
+
+We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
+to cover the rest of the university population.
+
+For each device instance, we sample from the appropriate distribution of embodied emissions,
+power draw, and duty cycles to obtain the devices characteristics.
+
+Combining power draw and duty cycle with the Swiss electricity mix, we compute annual operational
+emissions and add them to the embodied emissions scaled by the device type's expected lifetime.
+
+As mentioned, a possible future refinement will be to model the duty cycle as a function of time of day
+and incorporate finer-grained electricity mix data.
+
+\subsection{Cloud emissions}
+
+We model cloud emissions based on the assumptions and data described in
+Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
+more accurate data.
+
+\subsection{LLM emissions}
+
+We model LLM emissions by sampling from the per-capita daily query distribution and computing
+the carbon intensity of each query based on the average European electricity mix. While there 
+are many uncertainties about the level of LLM adoption, there appears to be reasonable consensus
+in the literature about the energy consumption of each query.
+
+\section{Missing data}
+
+The following data would improve the accuracy of our model (in roughly decreasing order of utility):
+\begin{itemize}
+    \item Campus data center energy use (e.g., at the granularity of monthly electricity bills).
+    \item Current campus data center workloads (to better model transition to the cloud).
+    \item Academic calendar (e.g., how many days per year is the university full vs operating
+        at lower levels).
+    \item Common or otherwise representative pieces of equipment (servers, routers, standard desktops, etc.).
+    \item Rough breakdown of UZH's population by field / role.
+\end{itemize}
 

### 2024-12-03 doc.tex
Update power estimates.

@@
 \author{Massimiliano Poletto}
-\date{\vspace{-1em}29.11.2024}
+\date{\vspace{-1em}29.11.2024\footnote{Last modified: \today}}
 
@@
 Monitor & 50 & 10 \\ \hline
-Conference display & 120 & 25 \\ \hline
+Conference display & 250 & 50 \\ \hline
 Printer/copier & 1000 & 200 \\ \hline

### 2024-12-18 it-emissions.tex
More meaningful filename
Doc improvements: explain model before explaining data, add appropriate caveats about stochastic approach.

--- 2024-12-03 doc.tex
@@
 \renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
-\newcommand{\assumption}[1]{{\bf #1}}
+\newcommand{\assumption}[1]{{#1}}
 %\usepackage{helvet} % Helvetica font for sans-serif
@@
 \author{Massimiliano Poletto}
-\date{\vspace{-1em}29.11.2024\footnote{Last modified: \today}}
+\date{\vspace{-1em}Last modified: 18.12.2024}
 
@@
 
+\section{Modeling approach}
+
+At a high level, we model IT emissions primarily bottom-up, based on device counts and
+estimates of their characteristics (embodied emissions, power utilization, and duty cycle).
+
+The model can operate in two modes: deterministic, where each device type has a fixed set of
+characteristics based on the mean of values we found in the literature, and stochastic, where
+we pick device characteristics from truncated normal distributions with the mean and standard
+deviation of the literature values. Of course, normality is a strong and probably incorrect
+assumption, but it is a simple initial way to model the uncertainty in our estimates.
+
+\subsection{Campus emissions}
+
+We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
+to cover the rest of the university population.
+
+For each device instance, we pick embodied emissions, power draw, and duty cycles either
+deterministically or by sampling from the appropriate distribution.
+
+Combining power draw and duty cycle estimates with ground-truth data about the Swiss electricity mix,
+we estimate annual operational emissions and add them to the embodied emissions scaled by the device
+type's expected lifetime.
+
+For on-campus data centers, for which presumably there are electricity bills or other measures of
+overall energy use, we would like to use that data to evaluate our model.
+
+A possible future refinement will be to model the duty cycle as a function of time of day
+and incorporate finer-grained electricity mix data.
+
+\subsection{Cloud emissions}
+
+We model cloud emissions based on the assumptions and data described in
+Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
+more accurate data.
+
+\subsection{LLM emissions}
+
+We model LLM emissions by sampling from the per-capita daily query distribution and computing
+the carbon intensity of each query based on the average European electricity mix. While there 
+are many uncertainties about the level of LLM adoption, there appears to be reasonable consensus
+in the literature about the energy consumption of each query.
+
 \section{Data sources and unknowns}
@@
 we assume that the entire university population has a laptop and a mobile phone.}
-
 We do not yet have any information about the breakdown of different functions / job roles, so
-\assumption{we assume that IT use is uniform across the university}. This assumption should be
-revisited.
+\assumption{we assume that IT use is uniform across the university}. Both of these assumptions
+should be revisited.
 
 \subsection{Embodied emissions}
-The estimation of embodied emissions introduces additional assumptions and uncertainties into
-the model. Our initial project proposal ignored embodied emissions altogether, but that leaves
-too big a gap.
 
-We did some research on the estimated embodied emissions of different types of devices.
+For several device types, especially end-user devices such as phones and laptops, embodied emissions
+comprise a majority of lifetime emissions. However, the exact amount of these emissions varies
+widely between different sources. Our approach has been to find several studies and industry references,
+and then estimate means and standard deviations from those that seemed most comprehensive and reliable
+(for example, Dell's Product Carbon Footprint reports about its servers).
 
@@
 A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
-\assumption{We model desktop supply chain emissions using a truncated normal distribution with those characteristics.}
+\assumption{We model desktop supply chain emissions using these statistics.}
 
@@
 1912 kg with $\sigma = 885$ kg.
-\assumption{We model server supply chain emissions using a truncated normal distribution corresponding
-to Dell servers.}
+\assumption{We model server supply chain emissions using the statistics corresponding to Dell servers.}
 
@@
 
-\assumption{We model phone supply chain emissions using a truncated normal distribution with mean
-50 kg CO$_2$e and $\sigma = 10$ kg.}
+\assumption{We model phone supply chain emissions using mean 50 kg CO$_2$e and $\sigma = 10$ kg.}
 
@@
 Based on a smattering of data sheets, primarily from Apple, Cisco, and Dell, we model
-power draw of different devices while under load as truncated normal distributions with the
-following initial parameters:
+power draw of different devices while under load using the following initial parameters:
 
@@
 
-Initially, we assume a distribution of duty cycles and power draw when idle for each device.
-As a next step, we may simulate power draw at different times of day to integrate with
+We assume a distribution of duty cycles and power draw when idle for each device.
+In the future, we may simulate power draw at different times of day to integrate with
 hourly electricity mix data, but errors in those assumptions may swamp the relatively small
@@
 
-\section{Modeling approach}
-
-At a high level, we model IT emissions primarily bottom-up, based on device counts and
-characteristics, and use a Monte Carlo approach to account for uncertainties.
-
-For on-campus data centers, we would like to use end-to-end data on electricity use (electric bills)
-to evaluate our model.
-
-\subsection{Campus emissions}
-
-We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
-to cover the rest of the university population.
-
-For each device instance, we sample from the appropriate distribution of embodied emissions,
-power draw, and duty cycles to obtain the devices characteristics.
-
-Combining power draw and duty cycle with the Swiss electricity mix, we compute annual operational
-emissions and add them to the embodied emissions scaled by the device type's expected lifetime.
-
-As mentioned, a possible future refinement will be to model the duty cycle as a function of time of day
-and incorporate finer-grained electricity mix data.
-
-\subsection{Cloud emissions}
-
-We model cloud emissions based on the assumptions and data described in
-Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
-more accurate data.
-
-\subsection{LLM emissions}
-
-We model LLM emissions by sampling from the per-capita daily query distribution and computing
-the carbon intensity of each query based on the average European electricity mix. While there 
-are many uncertainties about the level of LLM adoption, there appears to be reasonable consensus
-in the literature about the energy consumption of each query.
-
 \section{Missing data}

### 2025-01-29 it-emissions.tex
Various model improvements: - more accurate user counts - clearer labels - PUE for on-campus data centers and cloud - streamlined code for usage and embodied emissions - training overhead / embodied emissions for LLMs

@@
 
-By contrast, \textcite{wikiuzh} reports approximately 28,000 students and 10,000 staff
+By contrast, \textcite{uzhprofile} reports approximately 28,000 students and 7,300 staff
 at UZH. Evidently the spreadsheet only lists university assets, not personal devices.
@@
 \assumption{In modeling overall university IT emissions as well as potential LLM use,
-we assume that the entire university population has a laptop and a mobile phone.}
+we assume that 90\% of the university population, 32,000 people, have a laptop and a mobile phone.}
 We do not yet have any information about the breakdown of different functions / job roles, so
@@
 rapidly as model hardware and software improve, even though overall energy use is increasing due to
-increased utilization.
-
-Nevertheless, for the model we assume that queries consume 3 Wh each, and we model per-capita daily
-queries using a truncated normal distribution with mean 15 and $\sigma = 5$.
+increased utilization. Nevertheless, for the model we assume that queries consume 3 Wh each.
+We assume that on average every user will issue 15 queries per day.
 
@@
 average European electricity mix to compute carbon intensity.
+
+The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
+However, \textcite{devries2023growing} provides the following information about ChatGPT:
+\begin{itemize}
+    \item Operational power consumption of 500 MWh / day.
+    \item Approximately 1300 MWh consumed in training.
+    \item Approximately 4000 servers.
+\end{itemize}
+
+Assuming a relatively clean grid at 200 g CO$_2$e / kWh (a low estimate), and that each server has embodied emissions of 4 t CO$_2$e (a high estimate),
+we obtain:
+\begin{itemize}
+    \item 100 t CO$_2$e / day operational emissions.
+    \item 260 t CO$_2$e training emissions.
+    \item 16 kt CO$_2$e total embodied emissions.
+\end{itemize}
+
+After one year of operation, operational emissions are approximately 36 kt CO$_2$e, whereas training and embodied emissions are about 16 kt CO$_2$e,
+so operational emissions account for $\approx 70\%$ of the total. With less conservative assumptions (dirtier grid, lower embodied emissions),
+the proportion of operational emissions increases further.
+
+As a result, when modeling overall LLM emissions, we multiply estimated operational emissions by 1.4 to account for training and embodied emissions.
+We acknowledge that this is, at best, a rough estimate based on just one LLM.
 

### 2025-06-22 doc/paper.tex
Skeleton for paper submission

--- 2024-12-03 doc.tex
@@
 \addbibresource{ref.bib}
-\title{UZH IT \& AI Emissions Analysis\vspace{-1em}}
+\title{Modeling contributions to the IT-related emissions of a university}
 \author{Massimiliano Poletto}
-\date{\vspace{-1em}29.11.2024\footnote{Last modified: \today}}
 
@@
 \maketitle
+
+\section{Abstract}
+
 \section{Introduction}
 
-We are trying to answer two questions:
-\begin{itemize}
-    \item Are there emissions savings in moving UZH's IT infrastructure to the cloud?
-    \item What is the impact of increased AI use (and, in particular, of LLMs) on UZH's emissions?
-\end{itemize}
+\section{Modeling methodology}
 
-To this end, we need some UZH-specific information:
-\begin{enumerate}
-    \item The size of UZH's student and staff population and a rough breakdown of
-        what kind of work they do (to estimate their use of IT and AI).
-    \item Information about UZH's current IT infrastructure (number of servers,
-        classroom equipment, etc.) and the energy use (electricity bills) of any
-        campus data centers.
-    \item An estimate of the cloud compute capacity required to host UZH's current IT requirements.
-\end{enumerate}
+\section{Results}
 
-In addition, we need information about the possible location of cloud data centers and
-LLM deployments, the embodied emissions of different types of IT equipment, and the carbon intensity
-of electricity in different regions.
+\section{Related work}
 
-We describe our modeling approach and our assumptions and uncertainties regarding these factors.
+\section{Conclusion}
 
-\section{Data sources and unknowns}
+\section{References}
 
-\subsection{UZH population}
-
-We have access to a spreadsheet (\texttt{2024-11-13\_Berechnung\_THG\_IKT\_2024.xls}) with
-counts and emissions factors for various types of IT equipment at UZH. It reports approximately
-10,000 laptops and monitors, 1,000 servers, 2,400 mobile phones,  150 classroom PCs, and a few
-hundred printers, copiers, and projectors.
-
-By contrast, \textcite{wikiuzh} reports approximately 28,000 students and 10,000 staff
-at UZH. Evidently the spreadsheet only lists university assets, not personal devices.
-
-\assumption{In modeling overall university IT emissions as well as potential LLM use,
-we assume that the entire university population has a laptop and a mobile phone.}
-
-We do not yet have any information about the breakdown of different functions / job roles, so
-\assumption{we assume that IT use is uniform across the university}. This assumption should be
-revisited.
-
-\subsection{Embodied emissions}
-The estimation of embodied emissions introduces additional assumptions and uncertainties into
-the model. Our initial project proposal ignored embodied emissions altogether, but that leaves
-too big a gap.
-
-We did some research on the estimated embodied emissions of different types of devices.
-
-\subsubsection{Laptops}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-121 & \textcite{ecoinvent}  \\ \hline
-104 & \textcite{teehan2013} (low) \\ \hline
-338 & \textcite{teehan2013} (high) \\ \hline
-244 & \textcite{rarecoil} \\ \hline
-202 & \textcite{unctadder2024} \\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:laptops}
-\end{center}
-
-The Ecoinvent numbers appear to be outliers on the low end.
-The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
-information for over 90 popular devices. Mean supply chain emissions are
-244 kg CO$_2$e with a standard deviation of 128 kg.
-\assumption{We model laptop supply chain emissions using a truncated normal distribution with those characteristics.}
-
-\subsubsection{Desktops}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-238 & \textcite{ecoinvent} \\ \hline
-303 & \textcite{teehan2013} \\ \hline
-403 & \textcite{unctadder2024} \\ \hline
-289 & \textcite{dellpcf} \\ \hline
-277 & \textcite{boavizta} \\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:desktops}
-\end{center}
-
-Most estimates are in the range of 250 - 400 kg CO$_2$e. The most detailed data comes from
-Dell's Product Carbon Footprints (\textcite{dellpcf}), which they publish for each of their products.
-A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
-\assumption{We model desktop supply chain emissions using a truncated normal distribution with those characteristics.}
-
-\subsubsection{Servers}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-383 & \textcite{teehan2013} \\ \hline
-1252 & \textcite{davy2021} (Dell) \\ \hline
-1912 & \textcite{davy2021} (EC2) \\ \hline
-899 & \textcite{boavizta} (medium server)\\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:servers}
-\end{center}
-
-\textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
-The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
-1912 kg with $\sigma = 885$ kg.
-\assumption{We model server supply chain emissions using a truncated normal distribution corresponding
-to Dell servers.}
-
-\subsubsection{Smartphones}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
-54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
-50 & \textcite{unctadder2024} \\ \hline
-50 & \textcite{lovehagen2023} \\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:phones}
-\end{center}
-
-\assumption{We model phone supply chain emissions using a truncated normal distribution with mean
-50 kg CO$_2$e and $\sigma = 10$ kg.}
-
-\subsubsection{Other devices}
-
-We found less data for other types of devices. For now, we model them with averages based
-on a handful of sources.
-
-\begin{description}
-    \item[Computer monitors] 344 kg CO$_2$e (\textcite{teehan2013}, \textcite{dellpcf})
-    \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
-    \item[Printer/copier stations] 1167 kg CO$_2$e (\textcite{ecoinvent})
-    \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
-      and have no information about UZH network architecture. The vast majority ($80-95\%$) of
-      lifecycle emissions of network gear stem from usage (\textcite{cisco2024}, \textcite{jacob2023}).
-      For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
-      (A typical data center rack contains 42 1U servers and one switch or router.)
-\end{description}
-
-\subsection{Operational emissions}
-
-For on-premise equipment, we would like to model the Swiss electricity mix (see below) and, ideally,
-daily variations in usage and electricity carbon intensity.
-Generic manufacturer estimates of lifetime usage-related emissions are therefore unsuitable.
-There are surprisingly few sources of data on device power consumption (as opposed to CO$_2$ emissions).
-Based on a smattering of data sheets, primarily from Apple, Cisco, and Dell, we model
-power draw of different devices while under load as truncated normal distributions with the
-following initial parameters:
-
-\begin{center}
-\begin{tabular}{|l|r|r|}
-\hline
-\textbf{Device} & \textbf{Mean (W)} & \textbf{$\sigma$ (W)} \\ \hline
-Laptop & 30 & 5 \\ \hline
-Desktop & 100 & 20 \\ \hline
-Server & 400 & 100 \\ \hline
-Phone & 5 & 2 \\ \hline
-Monitor & 50 & 10 \\ \hline
-Conference display & 250 & 50 \\ \hline
-Printer/copier & 1000 & 200 \\ \hline
-\end{tabular}
-\end{center}
-
-Initially, we assume a distribution of duty cycles and power draw when idle for each device.
-As a next step, we may simulate power draw at different times of day to integrate with
-hourly electricity mix data, but errors in those assumptions may swamp the relatively small
-hourly variations in Swiss electricity carbon intensity.
-
-\subsection{Cloud data centers}
-\label{sec:cloud_data_centers}
-
-The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
-infrastructure. The best resource we have been able to find is the Datavizta API (\textcite{boavizta}).
-
-Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as follows:
-\begin{itemize}
-    \item One-for-one replacement of one campus server with one Azure D8S\_v3 instance (2x Intel
-      Xeon, 8 cores, 32 GB RAM, 2TB SSD) (\cite{msftvms}).
-    \item Deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region .
-    \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 kg CO$_2$e for usage,
-      97 kg CO$_2$e for manufacturing.
-\end{itemize}
-
-\subsection{Large language models}
-
-Energy consumption of LLMs is being studied extensively (\textcite{budennyy2022eco2ai},
-\textcite{castano2023exploring},
-\textcite{devries2023growing},
-\textcite{gowda2024watt},
-\textcite{harding2024watts}, 
-\textcite{heguerte2023estimate},
-\textcite{luccioni2022estimating},
-\textcite{luccioni2023counting},
-\textcite{patterson2021carbon}, 
-\textcite{rodriguez2024evaluating},
-\textcite{tripp2024measuring}).
-
-The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request,
-or approximately 10x the energy of a traditional search engine query. This number appears to be decreasing
-rapidly as model hardware and software improve, even though overall energy use is increasing due to
-increased utilization.
-
-Nevertheless, for the model we assume that queries consume 3 Wh each, and we model per-capita daily
-queries using a truncated normal distribution with mean 15 and $\sigma = 5$.
-
-LLMs are distributed and the location that serves any particular request is unknown, so we use the 
-average European electricity mix to compute carbon intensity.
-
-\subsection{Electricity carbon intensity}
-
-The Swiss Federal Office for the Environment published an extensive report (\textcite{krebs2018umweltbilanz})
-on the environmental impact of electricity generation in Switzerland in 2018.
-
-However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
-we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
-at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
-hour block over the course of the year. This gives us twelve data points that describe average
-carbon intensity variation over the course of the day.
-
-\section{Modeling approach}
-
-At a high level, we model IT emissions primarily bottom-up, based on device counts and
-characteristics, and use a Monte Carlo approach to account for uncertainties.
-
-For on-campus data centers, we would like to use end-to-end data on electricity use (electric bills)
-to evaluate our model.
-
-\subsection{Campus emissions}
-
-We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
-to cover the rest of the university population.
-
-For each device instance, we sample from the appropriate distribution of embodied emissions,
-power draw, and duty cycles to obtain the devices characteristics.
-
-Combining power draw and duty cycle with the Swiss electricity mix, we compute annual operational
-emissions and add them to the embodied emissions scaled by the device type's expected lifetime.
-
-As mentioned, a possible future refinement will be to model the duty cycle as a function of time of day
-and incorporate finer-grained electricity mix data.
-
-\subsection{Cloud emissions}
-
-We model cloud emissions based on the assumptions and data described in
-Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
-more accurate data.
-
-\subsection{LLM emissions}
-
-We model LLM emissions by sampling from the per-capita daily query distribution and computing
-the carbon intensity of each query based on the average European electricity mix. While there 
-are many uncertainties about the level of LLM adoption, there appears to be reasonable consensus
-in the literature about the energy consumption of each query.
-
-\section{Missing data}
-
-The following data would improve the accuracy of our model (in roughly decreasing order of utility):
-\begin{itemize}
-    \item Campus data center energy use (e.g., at the granularity of monthly electricity bills).
-    \item Current campus data center workloads (to better model transition to the cloud).
-    \item Academic calendar (e.g., how many days per year is the university full vs operating
-        at lower levels).
-    \item Common or otherwise representative pieces of equipment (servers, routers, standard desktops, etc.).
-    \item Rough breakdown of UZH's population by field / role.
-\end{itemize}
-
-\section{Other considerations}
-
-\subsection{Other IT components}
-
-For now, we do not attempt to model changes in network infrastructure or utilization due to
-move to the cloud or increased LLM use. See \textcite{jacob2023} for an overview of
-network sustainability.
-
-\subsection{Potential AI benefits}
-
-We typically think of AI use as creating additional energy demands and GHG emissions.
-So far, this model does not include potential energy savings caused by AI, such as improvements
-to building energy management, campus logistics, etc.
-
-\textcite{tomlinson2024carbon} make the argument that tasks such as writing
-and illustrating can be done with fewer GHG emissions by AI than by humans. They compare the
-estimated emissions of an LLM performing the task to those of the average human, calculated
-by multiplying per-capita annual emissions by the time ($\approx 1$h) required to write or draw.
-Using this approach, AI is orders of magnitude cheaper than humans, whether one uses per-capita
-emissions from the US (15 t CO$_2$e/y) or India (1.9 t). Of course, this methodology focuses on
-a specific creative act: the LLM does not reduce the human's overall annual emissions.
-
-\printbibliography
 \end{document}

### 2025-07-31 doc/paper.tex
Abstract

@@
 \addbibresource{ref.bib}
-\title{Modeling contributions to the IT-related emissions of a university}
-\author{Massimiliano Poletto}
+\title{Modeling the impact of IT- and AI-related emissions in a university setting.}
+\author{Massimiliano Poletto, Nils Handler}
 
@@
 \section{Abstract}
+Information technology (IT) plays major role in society and it has a non-negligible carbon footprint: for example, at the University of Zürich, it accounted for over $6\%$ of total emissions in 2024.
+At the same time, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates.
+Their growth and relatively high energy requirements are expected to increase the share of global electricity consumption for data centers from $\approx 1\%$ in 2024 to $\approx 3\%$ by 2030.
+Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap.
+As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and break down the impact of IT and AI on the university's carbon footprint.
+In this paper we present a simple interactive emissions model that approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs.
+Our analysis shows that IT infrastructure's embodied emissions are a dominant component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $3-5\%$ of overall emissions.
 
 \section{Introduction}
+
+approximately $6.3\%$ of total emissions in 2024\cite{uzh:sustainability:report},
+
+artificial intelligence (AI) applications are being adopted at historically high rates exceeding those of the adoption of the personal computer and the internet\cite{bick:ai:adoption}.
 
@@
 
+\printbibliography
+
 \end{document}

### 2025-07-31 doc/paper.tex
Modeling, top results, initial merge of stuff from it-emissions.tex

--- 2025-01-29 it-emissions.tex
@@
 \usepackage{hyperref}
-\usepackage[backend=biber,style=authoryear,autocite=inline]{biblatex}
+\hypersetup{
+    hidelinks
+}
+\usepackage[style=numeric]{biblatex}
 \usepackage{titlesec}
@@
 \addbibresource{ref.bib}
-\title{UZH IT \& AI Emissions Analysis\vspace{-1em}}
-\author{Massimiliano Poletto}
-\date{\vspace{-1em}Last modified: 18.12.2024}
+\title{Modeling the impact of IT- and AI-related emissions in a university setting.}
+\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}
 
@@
 \maketitle
+
+\section{Abstract}
+Information technology (IT) plays major role in society and has a non-negligible carbon footprint: for example, at the University of Zürich, it accounted for over $6\%$ of total emissions in 2024.
+At the same time, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates.
+Their growth and relatively high energy requirements are expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030.
+Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap.
+As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and break down the impact of IT and AI on the university's carbon footprint.
+In this paper we present a simple interactive emissions model that approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs.
+Our analysis shows that IT infrastructure's embodied emissions are a dominant component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.
+
 \section{Introduction}
 
-We are trying to answer two questions:
-\begin{itemize}
-    \item Are there emissions savings in moving UZH's IT infrastructure to the cloud?
-    \item What is the impact of increased AI use (and, in particular, of LLMs) on UZH's emissions?
-\end{itemize}
+approximately $6.3\%$ of total emissions in 2024 \cite{uzh:sustainability:report},
 
-To this end, we need some UZH-specific information:
-\begin{enumerate}
-    \item The size of UZH's student and staff population and a rough breakdown of
-        what kind of work they do (to estimate their use of IT and AI).
-    \item Information about UZH's current IT infrastructure (number of servers,
-        classroom equipment, etc.) and the energy use (electricity bills) of any
-        campus data centers.
-    \item An estimate of the cloud compute capacity required to host UZH's current IT requirements.
-\end{enumerate}
+artificial intelligence (AI) applications are being adopted at historically high rates exceeding those of the adoption of the personal computer and the internet \cite{bick:ai:adoption}.
 
-In addition, we need information about the possible location of cloud data centers and
-LLM deployments, the embodied emissions of different types of IT equipment, and the carbon intensity
-of electricity in different regions.
+We consider only direct impacts of AI: the increased electricity consumption and resulting emissions.
+We omit indirect impacts of AI adoption, such as potential emissions reductions due to improved building management: at the moment they are too nebulous to predict.
 
-We describe our modeling approach and our assumptions and uncertainties regarding these factors.
+We attempt to model full lifecycle impact, including manufacture, operation, and disposal.
+We include not only university-owned devices (servers, classroom PCs, etc.), but also student and staff personal devices, since they represent the bulk of on-campus devices.
 
-\section{Modeling approach}
+Likewise, we attempt to model full life cycle impact for cloud infrastructure and for LLMs.
 
-At a high level, we model IT emissions primarily bottom-up, based on device counts and
-estimates of their characteristics (embodied emissions, power utilization, and duty cycle).
+\section{Modeling methodology}
 
-The model can operate in two modes: deterministic, where each device type has a fixed set of
-characteristics based on the mean of values we found in the literature, and stochastic, where
-we pick device characteristics from truncated normal distributions with the mean and standard
-deviation of the literature values. Of course, normality is a strong and probably incorrect
-assumption, but it is a simple initial way to model the uncertainty in our estimates.
+We model campus IT emissions bottom-up, estimating the number of different devices and their characteristics, including embodied emissions, maximum power use, duty cycle, and operational lifetime.
+The model can operate in a deterministic linear mode or in a stochastic mode that perturbs device parameters.
+
+Likewise we model cloud data centers bottom-up, using publicly available emissions data and assuming a one-for-one replacement of on-campus servers with cloud instances of a certain size.
+
+For LLMs, we estimate operational energy use (Wh per query) using recent public data.
+Non-operational emissions, including embodied emissions of the server fleet and the cost of training the models, are harder to estimate and attribute.
+Based on data available for ChatGPT, we assume that non-operational emissions are a $40\%$ overhead on top of operational emissions.
 
@@
 
-\section{Data sources and unknowns}
-
-\subsection{UZH population}
-
-We have access to a spreadsheet (\texttt{2024-11-13\_Berechnung\_THG\_IKT\_2024.xls}) with
-counts and emissions factors for various types of IT equipment at UZH. It reports approximately
-10,000 laptops and monitors, 1,000 servers, 2,400 mobile phones,  150 classroom PCs, and a few
-hundred printers, copiers, and projectors.
-
-By contrast, \textcite{uzhprofile} reports approximately 28,000 students and 7,300 staff
-at UZH. Evidently the spreadsheet only lists university assets, not personal devices.
-
-\assumption{In modeling overall university IT emissions as well as potential LLM use,
-we assume that 90\% of the university population, 32,000 people, have a laptop and a mobile phone.}
-We do not yet have any information about the breakdown of different functions / job roles, so
-\assumption{we assume that IT use is uniform across the university}. Both of these assumptions
-should be revisited.
-
 \subsection{Embodied emissions}
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-121 & \textcite{ecoinvent}  \\ \hline
-104 & \textcite{teehan2013} (low) \\ \hline
-338 & \textcite{teehan2013} (high) \\ \hline
-244 & \textcite{rarecoil} \\ \hline
-202 & \textcite{unctadder2024} \\ \hline
+121 & \cite{ecoinvent}  \\ \hline
+104 & \cite{teehan2013} (low) \\ \hline
+338 & \cite{teehan2013} (high) \\ \hline
+244 & \cite{rarecoil} \\ \hline
+202 & \cite{unctadder2024} \\ \hline
 \end{tabular}
@@
 The Ecoinvent numbers appear to be outliers on the low end.
-The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
+The most comprehensive data comes from \cite{rarecoil}, which lists manufacturer
 information for over 90 popular devices. Mean supply chain emissions are
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-238 & \textcite{ecoinvent} \\ \hline
-303 & \textcite{teehan2013} \\ \hline
-403 & \textcite{unctadder2024} \\ \hline
-289 & \textcite{dellpcf} \\ \hline
-277 & \textcite{boavizta} \\ \hline
+238 & \cite{ecoinvent} \\ \hline
+303 & \cite{teehan2013} \\ \hline
+403 & \cite{unctadder2024} \\ \hline
+289 & \cite{dellpcf} \\ \hline
+277 & \cite{boavizta} \\ \hline
 \end{tabular}
@@
 Most estimates are in the range of 250 - 400 kg CO$_2$e. The most detailed data comes from
-Dell's Product Carbon Footprints (\textcite{dellpcf}), which they publish for each of their products.
+Dell's Product Carbon Footprints (\cite{dellpcf}), which they publish for each of their products.
 A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-383 & \textcite{teehan2013} \\ \hline
-1252 & \textcite{davy2021} (Dell) \\ \hline
-1912 & \textcite{davy2021} (EC2) \\ \hline
-899 & \textcite{boavizta} (medium server)\\ \hline
+383 & \cite{teehan2013} \\ \hline
+1252 & \cite{davy2021} (Dell) \\ \hline
+1912 & \cite{davy2021} (EC2) \\ \hline
+899 & \cite{boavizta} (medium server)\\ \hline
 \end{tabular}
@@
 
-\textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
+\cite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
 The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
-54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
-50 & \textcite{unctadder2024} \\ \hline
-50 & \textcite{lovehagen2023} \\ \hline
+68 & \cite{googlepixel8} (Pixel 8)\\ \hline
+54 & \cite{appleiphone13} (iPhone 13)\\ \hline
+50 & \cite{unctadder2024} \\ \hline
+50 & \cite{lovehagen2023} \\ \hline
 \end{tabular}
@@
 \begin{description}
-    \item[Computer monitors] 344 kg CO$_2$e (\textcite{teehan2013}, \textcite{dellpcf})
+    \item[Computer monitors] 344 kg CO$_2$e (\cite{teehan2013}, \cite{dellpcf})
     \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
-    \item[Printer/copier stations] 1167 kg CO$_2$e (\textcite{ecoinvent})
+    \item[Printer/copier stations] 1167 kg CO$_2$e (\cite{ecoinvent})
     \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
       and have no information about UZH network architecture. The vast majority ($80-95\%$) of
-      lifecycle emissions of network gear stem from usage (\textcite{cisco2024}, \textcite{jacob2023}).
+      lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
       For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
@@
 The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
-infrastructure. The best resource we have been able to find is the Datavizta API (\textcite{boavizta}).
+infrastructure. The best resource we have been able to find is the Datavizta API (\cite{boavizta}).
 
@@
 
-Energy consumption of LLMs is being studied extensively (\textcite{budennyy2022eco2ai},
-\textcite{castano2023exploring},
-\textcite{devries2023growing},
-\textcite{gowda2024watt},
-\textcite{harding2024watts}, 
-\textcite{heguerte2023estimate},
-\textcite{luccioni2022estimating},
-\textcite{luccioni2023counting},
-\textcite{patterson2021carbon}, 
-\textcite{rodriguez2024evaluating},
-\textcite{tripp2024measuring}).
+Energy consumption of LLMs is being studied extensively (\cite{budennyy2022eco2ai},
+\cite{castano2023exploring},
+\cite{devries2023growing},
+\cite{gowda2024watt},
+\cite{harding2024watts},
+\cite{heguerte2023estimate},
+\cite{luccioni2022estimating},
+\cite{luccioni2023counting},
+\cite{patterson2021carbon},
+\cite{rodriguez2024evaluating},
+\cite{tripp2024measuring}).
 
@@
 The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
-However, \textcite{devries2023growing} provides the following information about ChatGPT:
+However, \cite{devries2023growing} provides the following information about ChatGPT:
 \begin{itemize}
@@
 
-The Swiss Federal Office for the Environment published an extensive report (\textcite{krebs2018umweltbilanz})
+The Swiss Federal Office for the Environment published an extensive report (\cite{krebs2018umweltbilanz})
 on the environmental impact of electricity generation in Switzerland in 2018.
 
-However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
+However, for the purposes of the model, we use data from \cite{electricitymaps}. Specifically,
 we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
@@
 
-\section{Missing data}
+\subsection{Limitations and validation}
 
-The following data would improve the accuracy of our model (in roughly decreasing order of utility):
+The biggest limitation is the modest availability of some ground truth data about UZH IT infrastructure.
+Specifically, we have only estimates of campus data center energy use, and limited information abut UZH research workloads (whether AI or not).
+We assume that many research AI workloads will be relatively similar in intensity to existing scientific computation workloads.
+We also have limited information (e.g., peak power consumption) about UZH use of external high-performance compute resources such as Alps/CSCS.
+
+Additionally, cloud vendors and AI companies are secretive about their infrastructure. We have relied on public datasets and research literature to estimate cloud and AI emissions.
+
+Despite these limitations, our bottom-up model aligns closely with two UZH data points.
+First, the model correctly estimates approximately $400$ kW of compute for central campus data centers.
+Second, for university-owned campus IT infrastructure, the model estimates $1.4$ kt CO$_2$e emissions per year, close to the $1.45$ kt reported in the 2024 university sustainability report \cite{uzh:sustainability:report}.
+
+\section{Results}
+
+Our analysis shows that manufacturing (“embodied”) emissions dominate UZH's IT carbon footprint, while operational impacts, including AI adoption, are relatively modest.
+
 \begin{itemize}
-    \item Campus data center energy use (e.g., at the granularity of monthly electricity bills).
-    \item Current campus data center workloads (to better model transition to the cloud).
-    \item Academic calendar (e.g., how many days per year is the university full vs operating
-        at lower levels).
-    \item Common or otherwise representative pieces of equipment (servers, routers, standard desktops, etc.).
-    \item Rough breakdown of UZH's population by field / role.
+
+\item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint.
+For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ t CO$_2$e annually.
+Usage-related emissions amount to about a tenth of that.
+
+\item Widespread adoption of LLMs and other frontier models should have moderate impact on total emissions.
+If everyone at UZH made heavy use of AI (15 queries daily), this would add approximately 220 t CO2e annually, of which 160 t directly attributable to queries.
+While this would almost double an individual's real-time energy use, the impact on overall emissions is relatively small.
+
+\item Embodied emissions are an important consideration for university-owned devices as well.
+If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ t CO$_2$e, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.
+
+\item Data center cooling and efficiency represents a potential savings opportunity.
+The largest operational emissions are from data center servers and cooling (320 t).
+UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders.
+Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ t CO$_2$e annually, though this estimate requires further validation.
+
 \end{itemize}
 
-\section{Other considerations}
+\section{Related work}
 
-\subsection{Other IT components}
+\section{Conclusion}
 
-For now, we do not attempt to model changes in network infrastructure or utilization due to
-move to the cloud or increased LLM use. See \textcite{jacob2023} for an overview of
-network sustainability.
-
-\subsection{Potential AI benefits}
-
-We typically think of AI use as creating additional energy demands and GHG emissions.
-So far, this model does not include potential energy savings caused by AI, such as improvements
-to building energy management, campus logistics, etc.
-
-\textcite{tomlinson2024carbon} make the argument that tasks such as writing
-and illustrating can be done with fewer GHG emissions by AI than by humans. They compare the
-estimated emissions of an LLM performing the task to those of the average human, calculated
-by multiplying per-capita annual emissions by the time ($\approx 1$h) required to write or draw.
-Using this approach, AI is orders of magnitude cheaper than humans, whether one uses per-capita
-emissions from the US (15 t CO$_2$e/y) or India (1.9 t). Of course, this methodology focuses on
-a specific creative act: the LLM does not reduce the human's overall annual emissions.
+\section{References}
 
 \printbibliography
+
 \end{document}

### 2025-07-31 doc/paper.tex
Checkpoint

@@
 \section{Abstract}
-Information technology (IT) plays major role in society and has a non-negligible carbon footprint: for example, at the University of Zürich, it accounted for over $6\%$ of total emissions in 2024.
+Information technology (IT) plays a major role in society and has non-negligible environmental impact, accounting for over 5\% of global electricity consumption and about 2\% of global carbon emissions.
 At the same time, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates.
@@
 Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap.
-As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and break down the impact of IT and AI on the university's carbon footprint.
-In this paper we present a simple interactive emissions model that approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs.
-Our analysis shows that IT infrastructure's embodied emissions are a dominant component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.
+As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and analyze the impact of IT and AI on the university's carbon footprint.
+In this paper we present a simple interactive model of IT emissions on a univeristy campus.
+The model approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs.
+Our analysis shows that the embodied emissions of IT infrastructure and personal devices are a major component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.
 
@@
 
+We do not consider potential benefits of AI.
+
 \section{Modeling methodology}
+
+\subsection{System boundaries}
+
+\subsection{Calculation approach}
 
@@
 Based on data available for ChatGPT, we assume that non-operational emissions are a $40\%$ overhead on top of operational emissions.
+\subsection{Parameters and assumptions}
+
+\section{Model details}
 
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-121 & \cite{ecoinvent}  \\ \hline
-104 & \cite{teehan2013} (low) \\ \hline
-338 & \cite{teehan2013} (high) \\ \hline
-244 & \cite{rarecoil} \\ \hline
-202 & \cite{unctadder2024} \\ \hline
+121 & \textcite{ecoinvent}  \\ \hline
+104 & \textcite{teehan2013} (low) \\ \hline
+338 & \textcite{teehan2013} (high) \\ \hline
+244 & \textcite{rarecoil} \\ \hline
+202 & \textcite{unctadder2024} \\ \hline
 \end{tabular}
@@
 The Ecoinvent numbers appear to be outliers on the low end.
-The most comprehensive data comes from \cite{rarecoil}, which lists manufacturer
+The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
 information for over 90 popular devices. Mean supply chain emissions are
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-238 & \cite{ecoinvent} \\ \hline
-303 & \cite{teehan2013} \\ \hline
-403 & \cite{unctadder2024} \\ \hline
-289 & \cite{dellpcf} \\ \hline
-277 & \cite{boavizta} \\ \hline
+238 & \textcite{ecoinvent} \\ \hline
+303 & \textcite{teehan2013} \\ \hline
+403 & \textcite{unctadder2024} \\ \hline
+289 & \textcite{dellpcf} \\ \hline
+277 & \textcite{boavizta} \\ \hline
 \end{tabular}
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-383 & \cite{teehan2013} \\ \hline
-1252 & \cite{davy2021} (Dell) \\ \hline
-1912 & \cite{davy2021} (EC2) \\ \hline
-899 & \cite{boavizta} (medium server)\\ \hline
+383 & \textcite{teehan2013} \\ \hline
+1252 & \textcite{davy2021} (Dell) \\ \hline
+1912 & \textcite{davy2021} (EC2) \\ \hline
+899 & \textcite{boavizta} (medium server)\\ \hline
 \end{tabular}
@@
 
-\cite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
+\textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
 The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
@@
 \textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
-68 & \cite{googlepixel8} (Pixel 8)\\ \hline
-54 & \cite{appleiphone13} (iPhone 13)\\ \hline
-50 & \cite{unctadder2024} \\ \hline
-50 & \cite{lovehagen2023} \\ \hline
+68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
+54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
+50 & \textcite{unctadder2024} \\ \hline
+50 & \textcite{lovehagen2023} \\ \hline
 \end{tabular}
@@
 \begin{description}
-    \item[Computer monitors] 344 kg CO$_2$e (\cite{teehan2013}, \cite{dellpcf})
+    \item[Computer monitors] 344 kg CO$_2$e (\textcite{teehan2013}, \textcite{dellpcf})
     \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
-    \item[Printer/copier stations] 1167 kg CO$_2$e (\cite{ecoinvent})
+    \item[Printer/copier stations] 1167 kg CO$_2$e (\textcite{ecoinvent})
     \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
       and have no information about UZH network architecture. The vast majority ($80-95\%$) of
-      lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
+      lifecycle emissions of network gear stem from usage (\textcite{cisco2024}, \textcite{jacob2023}).
       For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
@@
 The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
-However, \cite{devries2023growing} provides the following information about ChatGPT:
+However, \textcite{devries2023growing} provides the following information about ChatGPT:
 \begin{itemize}
@@
 
-However, for the purposes of the model, we use data from \cite{electricitymaps}. Specifically,
+However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
 we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
@@
 
+\section{Recommendations}
+
+Do not panic about LLMs. Their emissions comprise a small fraction of total IT emissions, they will get more efficient, and attempting to curb their use is likely to be a losing battle.
+Invest in infrastructure for the long term.
+Run awareness campaigns and adapt IT policies so that personal devices and university IT systems are used for as long as possible.
+Amortize manufacturing emissions over the longest period possible.
+Improve the cooling efficiency of university data centers, or engage cloud vendors and academic compute centers (such as CSCS) to validate the potential benefits of a migration to cloud.
+
 \section{Related work}
@@
 
-\section{References}
-
 \printbibliography

### 2025-08-01 doc/paper.tex
Checkpoint

@@
 
-approximately $6.3\%$ of total emissions in 2024 \cite{uzh:sustainability:report},
+Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions (estimates range from 1.5 to 4\%) \cite{wb:itu:ict}.
+At the University of Zürich, which with over 38,000 students and staff is the largest Swiss university, IT accounted for about 1.5 Mt CO$_2$e of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024.\cite{uzh:sustainability:report}.
+The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular generative AI and large language models (LLMs).
 
-artificial intelligence (AI) applications are being adopted at historically high rates exceeding those of the adoption of the personal computer and the internet \cite{bick:ai:adoption}.
+AI applications are being adopted at historically high rates, more quickly than the personal computer or the internet were adopted in the past\cite{bick:ai:adoption}.
+Enterprise AI adoption has accelerated, with generative AI tools seeing widespread deployment across industries within months of their release \cite{mckinsey2024ai}.
+iea:ai:energy
 

### 2025-08-02 doc/paper.tex
Checkpoint
Intro

--- 2025-07-31 doc/paper.tex
@@
 \addbibresource{ref.bib}
-\title{Modeling the impact of IT- and AI-related emissions in a university setting.}
+\title{Modeling the impact of IT- and AI-related emissions in a university setting}
 \author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}
@@
 Information technology (IT) plays a major role in society and has non-negligible environmental impact, accounting for over 5\% of global electricity consumption and about 2\% of global carbon emissions.
-At the same time, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates.
+Moreover, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates.
 Their growth and relatively high energy requirements are expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030.
@@
 As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and analyze the impact of IT and AI on the university's carbon footprint.
-In this paper we present a simple interactive model of IT emissions on a univeristy campus.
+We present a simple interactive model of IT emissions on a univeristy campus.
 The model approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs.
@@
 
-approximately $6.3\%$ of total emissions in 2024 \cite{uzh:sustainability:report},
+Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}.
+At the University of Zürich, which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt CO$_2$e of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
+The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).
 
-artificial intelligence (AI) applications are being adopted at historically high rates exceeding those of the adoption of the personal computer and the internet \cite{bick:ai:adoption}.
+AI applications are being adopted at historically high rates, more quickly than the personal computer or the internet were adopted in the past~\cite{bick:ai:adoption}.
+Enterprise AI adoption has accelerated, with generative AI tools seeing widespread deployment across industries within months of their release~\cite{mckinsey2024ai}.
+In July 2025, ChatGPT received 2.5 billion queries per day, up from 1 billion in December 2024, and within an order of magnitude of Google's approximately 15 billion queries per day~\cite{techcrunch:chatgpt}.
+AI growth is currently projected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030~\cite{iea:ai:energy}.
 
-We consider only direct impacts of AI: the increased electricity consumption and resulting emissions.
-We omit indirect impacts of AI adoption, such as potential emissions reductions due to improved building management: at the moment they are too nebulous to predict.
+The University of Zürich has set itself the target of becoming climate-neutral by 2030 within the categories of energy, transportation, food, and waste, with additional goals in 2035 and beyond~\cite{uzh:sustainability:report}.
+Given the current sizable contribution of IT to the university's emissions, and the potential for growth through AI adoption in the coming years, we felt it was necessary to understand IT emissions and their growth trajectories in more detail.
 
-We attempt to model full lifecycle impact, including manufacture, operation, and disposal.
-We include not only university-owned devices (servers, classroom PCs, etc.), but also student and staff personal devices, since they represent the bulk of on-campus devices.
+This paper presents a simple interactive model of IT emissions on a university campus.
+We developed it in the specific context of the University of Zürich, but believe that it is parameterizable and general enough to apply to other similar university campuses or large corporate facilities.
 
-Likewise, we attempt to model full life cycle impact for cloud infrastructure and for LLMs.
+We seek to answer three broad questions:
+\begin{itemize}
+  \item What are the main contributors to the IT emissions of a university?
+  \item Could migrating some on-campus IT services to commercial cloud providers decrease the overall campus footprint?
+  \item What is the likely impact of the surge in AI adoption? In particular, will the emissions impact of AI be so large as to make other sources of IT emissions negligible?
+\end{itemize}
 
-We do not consider potential benefits of AI.
+Our model sets relatively broad system boundaries.
+It takes into account university-owned equipment but also university services hosted by commercial cloud providers and university use of commercial generative AI products.
+We have also parameterized it to account for student devices not owned by the university, since they are key to student work and constitute the bulk of the university's end-user devices.
+We attempt to model devices' full lifecycle impact, including manufacture and disposal.
+
+Experts rightly point to the many potential benefits of AI in the areas of climate change mitigation and adapation~\cite{climate:change:ai}.
+On a university campus, these might include more efficient building heating and cooling, improved transportation logistics, etc.
+However, those benefits are still to come and difficult to predict.
+In this paper, we consider only the direct impacts of AI in terms of increased electricity consumption and the resulting emissions.
+
+The rest of this paper is organized as follows.
+Section~\ref{sec:model:methodology} presents the general modeling framework, including system boundaries and assumptions.
+Section~\ref{sec:model:details} justifies some of the specific parameter values used in the model.
+Section~\ref{sec:results} describes results and explores different scenarios and assumptions.
+Section~\ref{sec:recommendations} offers recommendations to university IT administrators.
+Section~\ref{sec:related:work} describes related work in this area.
 
 \section{Modeling methodology}
+\label{sec:model:methodology}
 
@@
 Based on data available for ChatGPT, we assume that non-operational emissions are a $40\%$ overhead on top of operational emissions.
+
 \subsection{Parameters and assumptions}
@@
 \section{Model details}
+\label{sec:model:details}
 
@@
 \section{Results}
+\label{sec:results}
 
@@
 \section{Recommendations}
+\label{sec:recommendations}
 
@@
 \section{Related work}
+\label{sec:related:work}
 

### 2025-08-02 doc/paper.tex
Methodology 1

@@
 \newcommand{\assumption}[1]{{#1}}
+\newcommand{\eg}{{\em e.g.}}
 %\usepackage{helvet} % Helvetica font for sans-serif
@@
 Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}.
-At the University of Zürich, which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt CO$_2$e of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
+At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt CO$_2$e (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
 The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).
@@
 However, those benefits are still to come and difficult to predict.
-In this paper, we consider only the direct impacts of AI in terms of increased electricity consumption and the resulting emissions.
+In this paper, we consider only the {\em direct} impacts of AI in terms of increased electricity consumption and the resulting emissions.
 
@@
 
+Our aim is to model energy use from unveristy IT and AI broadly, without restricting ourselves to university-owned devices, specific IT departments, etc.
+At the same time, we want to be able to distinguish the contributions of different factors, such as student-owned devices or LLMs, and to compare the model's output against existing university data points (energy bills, etc.).
+
 \subsection{System boundaries}
+
+We define the boundaries of our model as follows.
+
+\subsubsection*{Population}
+
+We include IT-related activity from all members of the UZH community: faculty, staff, and students.
+
+\subsubsection*{Devices}
+
+Student-owned devices, such as laptops and smartphones, are in scope, since they are the primary means by which students access university IT services and do their coursework.
+Given the duration of studies, we include personal devices' embodied emissions in the model.
+
+University-owned devices are in scope.
+We have accurate numbers for staff workstations, classroom computers, network access points, and audiovisual devices managed by the central IT department.
+We can only estimate other devices, such as computers managed by individual research groups or networking equipment.
+As before, we include these devices' embodied emissions in the model.
+
+\subsubsection*{Cloud and supercomputing facilities}
+
+UZH at present does not make significant use of commercial cloud compute, but the model allows simulating the shift of some fraction of the server load to cloud VMs.
+
+Due to a lack of data we do not attempt to model emissions from application-level cloud services such as Microsoft 365 or Google Workspace.
+This omission leads to undercounting total IT energy use but to overestimating the potential relative impact of AI.
+
+UZH faculties do make use of shared off-campus supercomputing facilities, primarily the Swiss National Supercomputing Center, CSCS~\cite{cscs}.
+We estimate the impact of these workloads based on available gross peak load numbers.
+Due to lack of data, we do not include the embodied emissions of these facilities.
+
+\subsubsection*{Large language models}
+
+We estimate inference-time emissions of current large language models based on the literature.
+It is harder to come by good estimates of training-time emissions and embodied emissions of the hardware, but we attempt to do so nonetheless in order to not risk undercounting the impact of LLMs.
+
+\subsubsection*{Summary}
+
+This table summarizes what is in scope in the current model. The main gap is emissions of cloud services such as Microsoft 365 or Goole Workspace.
+
+\begin{center}
+  \begin{tabular}{|l|c|c|}
+  \hline
+  \textbf{Device or service} & \textbf{Operation} & \textbf{Lifecycle} \\ \hline
+  Student laptops \& phones & yes & yes \\
+  University workstations, access points, etc. & yes & yes \\
+  Cloud servers & yes & yes \\
+  Cloud services & no & no \\
+  Supercomputing facilities & yes & no \\
+  LLMs & yes & yes \\ \hline
+  \end{tabular}
+  \label{tab:scopes}
+  \end{center}
 
@@
 
-We model campus IT emissions bottom-up, estimating the number of different devices and their characteristics, including embodied emissions, maximum power use, duty cycle, and operational lifetime.
+\subsubsection*{Devices}
+
+We model campus device emissions bottom-up.
+We first identify categories of devices (phones, laptops, workstations, monitors, routers, etc.).
+For each device type, we research a range of common models, then use manufacturer data sheets or other available documentation to identify operational power draw (in watts) as well as embodied emissions related to manufacture and disposal (in kg CO$_2$e) per device.
+Based on this data, we model the power draw and embodied emissions of an ``exemplar'' device of each type.
+
+For each device type we also estimate (1) a count (\eg, a laptop for almost every student and staff), (2) an estimated annual duty cycle (\eg, laptops used 8 hours per day, 5 days per week, 40 weeks per year, etc.), and (3) a useful lifetime (in years).
+
+Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (kg CO$_2$e).
+
+We then estimate the local grid's emissions intensity (g CO$_2$e / kWh) to convert the energy use to emissions and obtain a combined total annual emissions number.
+
+In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution. However, we have no good data on the actual shape of such a distribution,
+
+\subsubsection{Cloud and supercomputing facilities}
+
+Given
+
+estimating the number of different devices and their characteristics, including embodied emissions, maximum power use, duty cycle, and operational lifetime.
 The model can operate in a deterministic linear mode or in a stochastic mode that perturbs device parameters.

### 2025-08-02 doc/paper.tex
More parameters

@@
 
-\section{Abstract}
+\section*{Abstract}
 Information technology (IT) plays a major role in society and has non-negligible environmental impact, accounting for over 5\% of global electricity consumption and about 2\% of global carbon emissions.
-Moreover, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates.
-Their growth and relatively high energy requirements are expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030.
+Moreover, artificial intelligence (AI) applications, especially large language models (LLMs), are being adopted at historically high rates.
+Their growth is expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030.
 Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap.
 As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and analyze the impact of IT and AI on the university's carbon footprint.
-We present a simple interactive model of IT emissions on a univeristy campus.
+We present a simple interactive model of IT emissions on a university campus.
 The model approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs.
+The model is intended to provide university decision makers with emissions estimates under different scenarios, with sufficient accuracy to support effective decision-making.
 Our analysis shows that the embodied emissions of IT infrastructure and personal devices are a major component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.
@@
 
-Our aim is to model energy use from unveristy IT and AI broadly, without restricting ourselves to university-owned devices, specific IT departments, etc.
+Our aim is to model energy use from university IT and AI broadly, without restricting ourselves to university-owned devices, specific IT departments, etc.
 At the same time, we want to be able to distinguish the contributions of different factors, such as student-owned devices or LLMs, and to compare the model's output against existing university data points (energy bills, etc.).
@@
 
-For each device type we also estimate (1) a count (\eg, a laptop for almost every student and staff), (2) an estimated annual duty cycle (\eg, laptops used 8 hours per day, 5 days per week, 40 weeks per year, etc.), and (3) a useful lifetime (in years).
+For each device type we also estimate (1) a count (\eg, a laptop for every student and staff), (2) an estimated annual duty cycle (\eg, laptops used 8 hours per day, 200 days per year), and (3) a useful lifetime (in years).
 
@@
 We then estimate the local grid's emissions intensity (g CO$_2$e / kWh) to convert the energy use to emissions and obtain a combined total annual emissions number.
+We obtain emissions intensity data from Electricity Maps~\cite{electricitymaps}.
 
-In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution. However, we have no good data on the actual shape of such a distribution,
+In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution.
+However, we have no good data on the precise form of such a distribution, so we limit our analysis to the linear mode.
 
-\subsubsection{Cloud and supercomputing facilities}
+\subsubsection*{Cloud and supercomputing facilities}
 
-Given
+We attempt to model a hypothetical shift of some workloads from servers in an on-premise data center to a commercial cloud. The parameters under consideration include:
 
-estimating the number of different devices and their characteristics, including embodied emissions, maximum power use, duty cycle, and operational lifetime.
-The model can operate in a deterministic linear mode or in a stochastic mode that perturbs device parameters.
+\begin{itemize}
+  \item On-premise power usage effectiveness (PUE), the ratio of total data center power (including cooling) to power used for computation.
+  \item Cloud data center PUE and grid emissions intensity (since the cloud data center may be located in a different region than the on-premise data center).
+  \item An average conversion ratio of on-premise servers to cloud servers.
+\end{itemize}
 
-Likewise we model cloud data centers bottom-up, using publicly available emissions data and assuming a one-for-one replacement of on-campus servers with cloud instances of a certain size.
+We measure the impact of cloud servers in terms of annual operational emissions and amortized emissions due to manufacturing and disposal, in units of kg CO$_2$e / year.
 
-For LLMs, we estimate operational energy use (Wh per query) using recent public data.
-Non-operational emissions, including embodied emissions of the server fleet and the cost of training the models, are harder to estimate and attribute.
-Based on data available for ChatGPT, we assume that non-operational emissions are a $40\%$ overhead on top of operational emissions.
+We do not attempt to model application-level cloud services.
+
+For supercomputing facilities (in particular, CSCS) we rely on annual statistics of total power consumption for peak compute load and make assumptions about duty cycle.
+We do not have reliable data about the embodied emissions of the facility, but suspect that the fraction of these emissions that is attributable to UZH use is negligible relative to all other UZH emissions.
+
+\subsubsection*{Large language models}
+
+For LLMs, we estimate operational energy use (Wh per query) using current estimates in the literature of state-of-the-art models~\cite{devries2023growing,epoch2025howmuchenergydoeschatgptuse}.
+
+We consider the training overhead and the embodied emissions of the servers and GPUs that the model runs as a fixed cost that is amortized across all queries to a model.
+As a result, in practice, this overhead is constantly shrinking.
+We incorporate it into the model as a configurable percentage overhead of the operational energy use.
 
 \subsection{Parameters and assumptions}
+
+The model's outputs are affected by several factors, and where we were unable to find good data we made assumptions that we considered reasonable.
+All of these parameters and assumptions are parameterizable by users of the model.
+We list these sources of uncertainty below.
+
+\begin{description}
+  \item[Cloud data center PUE] As of 2025, industry average power usage effectiveness is 1.56, with Google averaging 1.09 across its fleet~\cite{google:datacenter:efficiency} and Microsoft ranging from 1.12 to 1.30~\cite{microsoft:datacenter:efficiency}.
+We choose a PUE of 1.2 as representative of a state-of-the-art hyperscaler data center.
+  \item[On-premise data center PUE] Based on limited historical power data, we assume an on-premise PUE of 2.0, attributable to inefficient cooling.
+  \item[]  
+\end{description}
 

### 2025-08-02 doc/paper.tex
More parameters

@@
   \item[On-premise data center PUE] Based on limited historical power data, we assume an on-premise PUE of 2.0, attributable to inefficient cooling.
+  \item[Duty cycle] We make some unsubstantiated but plausible assumptions about the number of days that different devices are used (\eg, end-user devices are not used every day) and their duty cycle throughut the day.
+  \item[Electricity emissions intensity for on-premise data centers] For Switzerland we compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps} ($\sim 50$ g / kWh on average).
+  \item[Electricity emissions intensity for cloud data centers and LLMs] We use the European Union's average emissions intensity of electricity generation in 2023, 207 g CO$_2$e / kWh~\cite{eea:emissions:2025}.
   \item[] 

### 2025-08-03 doc/paper.tex
Checkpoint

--- 2025-08-02 doc/paper.tex
@@
 \documentclass[11pt]{article}
-\usepackage[margin=0.7in]{geometry}
+\usepackage[margin=0.9in]{geometry}
 \usepackage[T1]{fontenc}
@@
 \newcommand{\eg}{{\em e.g.}}
+\newcommand{\coe}{CO$_2$e}
+\newcommand{\gcoe}{g \coe}
+\newcommand{\kgcoe}{k\gcoe}
+\newcommand{\gcoekwh}{\gcoe\ / kWh}
 %\usepackage{helvet} % Helvetica font for sans-serif
@@
 Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}.
-At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt CO$_2$e (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
+At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt \coe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
 The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).
@@
 
-\subsubsection*{Devices}
+\subsubsection*{On-campus devices}
 
@@
 We have accurate numbers for staff workstations, classroom computers, network access points, and audiovisual devices managed by the central IT department.
-We can only estimate other devices, such as computers managed by individual research groups or networking equipment.
+We can only estimate other devices, such as networking equipment or computers managed by individual research groups.
 As before, we include these devices' embodied emissions in the model.
@@
 
-Due to a lack of data we do not attempt to model emissions from application-level cloud services such as Microsoft 365 or Google Workspace.
+Due to lack of data, we do not attempt to model emissions from application-level cloud services such as Microsoft 365 or Google Workspace.
 This omission leads to undercounting total IT energy use but to overestimating the potential relative impact of AI.
@@
 UZH faculties do make use of shared off-campus supercomputing facilities, primarily the Swiss National Supercomputing Center, CSCS~\cite{cscs}.
-We estimate the impact of these workloads based on available gross peak load numbers.
-Due to lack of data, we do not include the embodied emissions of these facilities.
+We estimate the operational impact of these workloads, but, again due to lack of data, we do not include embodied emissions.
+Again, this undercounts total university IT energy use but overestimates the potential relative impact of AI.
 
@@
 We estimate inference-time emissions of current large language models based on the literature.
-It is harder to come by good estimates of training-time emissions and embodied emissions of the hardware, but we attempt to do so nonetheless in order to not risk undercounting the impact of LLMs.
+We also include training-time emissions and embodied emissions of language models (thouse these are harder to estimate and we err on the side of overestimating them).
 
@@
 
-\subsection{Calculation approach}
+\subsection{Accounting approach}
 
-\subsubsection*{Devices}
+\subsubsection*{On-campus devices}
 
@@
 We first identify categories of devices (phones, laptops, workstations, monitors, routers, etc.).
-For each device type, we research a range of common models, then use manufacturer data sheets or other available documentation to identify operational power draw (in watts) as well as embodied emissions related to manufacture and disposal (in kg CO$_2$e) per device.
+For each device type, we research a range of common models, then use manufacturer data sheets or other available documentation to identify operational power draw (in watts) as well as embodied emissions related to manufacture and disposal (in \kgcoe) per device.
 Based on this data, we model the power draw and embodied emissions of an ``exemplar'' device of each type.
 
-For each device type we also estimate (1) a count (\eg, a laptop for every student and staff), (2) an estimated annual duty cycle (\eg, laptops used 8 hours per day, 200 days per year), and (3) a useful lifetime (in years).
+For each device type we also estimate (1) a count (\eg, a laptop for every student and staff), (2) an estimated duty cycle (\eg, laptops used 8 hours per day, 200 days per year), and (3) a useful lifetime (in years).
 
-Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (kg CO$_2$e).
+Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (\kgcoe).
 
-We then estimate the local grid's emissions intensity (g CO$_2$e / kWh) to convert the energy use to emissions and obtain a combined total annual emissions number.
+We then estimate the local grid's emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number.
 We obtain emissions intensity data from Electricity Maps~\cite{electricitymaps}.
@@
 In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution.
-However, we have no good data on the precise form of such a distribution, so we limit our analysis to the linear mode.
+However, we have insufficient data on the precise form of such a distribution, so we limit our analysis to the linear mode.
 
@@
 
-We measure the impact of cloud servers in terms of annual operational emissions and amortized emissions due to manufacturing and disposal, in units of kg CO$_2$e / year.
+We measure the impact of cloud servers in terms of annual operational emissions and amortized emissions due to manufacturing and disposal, in units of \kgcoe\ / year.
 
-We do not attempt to model application-level cloud services.
+As mentioned earlier, we do not attempt to model application-level cloud services.
 
@@
 We choose a PUE of 1.2 as representative of a state-of-the-art hyperscaler data center.
-  \item[On-premise data center PUE] Based on limited historical power data, we assume an on-premise PUE of 2.0, attributable to inefficient cooling.
-  \item[]  
+  \item[On-premise data center PUE] Based on limited UZH historical power data, we assume an on-premise PUE of 2.0, attributable to inefficient cooling.
+  \item[Device duty cycle] We make unsupported but plausible assumptions about the number of days that different devices are used (\eg, end-user devices are not used every day) and their duty cycle throughut the day.
+  \item[Electricity emissions intensity for on-premise data centers] For Switzerland, we can compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps}.
+Emissions vary from a high of 64 \gcoekwh\ in the middle of the night to a low of 45 \gcoekwh\ in the middle of the day.
+This variation is sufficiently large that we refine the model's device duty cycle to four six-hour periods over the course of the day.
+It allows us to specify, for example, that user-devices are mostly used during the day whereas servers run around the clock.
+  \item[Electricity emissions intensity for cloud data centers and LLMs] These facilities are unlikely to be located in Switzerland, and Switzerland's grid is cleaner than average.
+We use the European Union's average emissions intensity of electricity generation in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
+  \item[LLM usage] We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).
 \end{description}
@@
 
-\subsection{Campus emissions}
+The previous section describes our overall modeling approach, which we believe is broadly applicable to university and other large campus settings.
+This section presents the reasoning behind specific default values we employ in the UZH evaluation.
+
+\subsection{On-campus devices}
+
+\subsubsection*{Number of devices}
+
+We have an authoritative census of devices (workstations, servers, audiovisual equipment) managed by central IT, as follows.
+
+\begin{center}
+  \begin{tabular}{|l|r|r|}
+  \hline
+  \textbf{Device type} & \textbf{Count} & \textbf{Life (y)}\\ \hline
+  Laptops & 10,440 & 4 \\
+  Monitors & 10,400 & 6 \\
+  Servers & 1,121 & 5 \\
+  Printers & 333 & 6 \\
+  Mobile phones & 2,439 & 3 \\
+  A/V devices & 431 & 6 \\ \hline
+  \end{tabular}
+  \label{tab:oncampusdevices}
+\end{center}
+
+On top of this, we assume that
+
+\subsubsection*{Operational emissions}
+
+\subsubsection*{Embodied emissions}
+
+We obtain
+
+\subsection{Cloud and supercomputing facilities}
+
+\subsection{LLMs}
 
@@
 \textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
+\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
 121 & \textcite{ecoinvent}  \\ \hline
@@
 information for over 90 popular devices. Mean supply chain emissions are
-244 kg CO$_2$e with a standard deviation of 128 kg.
+244 \kgcoe\ with a standard deviation of 128 kg.
 \assumption{We model laptop supply chain emissions using a truncated normal distribution with those characteristics.}
@@
 \textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
+\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
 238 & \textcite{ecoinvent} \\ \hline
@@
 
-Most estimates are in the range of 250 - 400 kg CO$_2$e. The most detailed data comes from
+Most estimates are in the range of 250 - 400 \kgcoe. The most detailed data comes from
 Dell's Product Carbon Footprints (\cite{dellpcf}), which they publish for each of their products.
-A sample of 15 desktops has mean supply chain emissions of 289 kg CO$_2$e with $\sigma = 80$ kg.
+A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg.
 \assumption{We model desktop supply chain emissions using these statistics.}
@@
 \textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
+\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
 383 & \textcite{teehan2013} \\ \hline
@@
 \textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
-The former have mean supply chain emissions of 1252 kg CO$_2$e with $\sigma = 330$ kg, the latter
+The former have mean supply chain emissions of 1252 \kgcoe\ with $\sigma = 330$ kg, the latter
 1912 kg with $\sigma = 885$ kg.
@@
 \textbf{Production \& Disposal} & \\
-\textbf{(kg CO$_2$e)} & \textbf{Source} \\ \hline
+\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
 68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
@@
 
-\assumption{We model phone supply chain emissions using mean 50 kg CO$_2$e and $\sigma = 10$ kg.}
+\assumption{We model phone supply chain emissions using mean 50 \kgcoe\ and $\sigma = 10$ kg.}
 
@@
 \begin{description}
-    \item[Computer monitors] 344 kg CO$_2$e (\textcite{teehan2013}, \textcite{dellpcf})
-    \item[Conference room displays] 753 kg CO$_2$e (scaling monitor by $(40''/27'')^2$)
-    \item[Printer/copier stations] 1167 kg CO$_2$e (\textcite{ecoinvent})
+    \item[Computer monitors] 344 \kgcoe\ (\textcite{teehan2013}, \textcite{dellpcf})
+    \item[Conference room displays] 753 \kgcoe\ (scaling monitor by $(40''/27'')^2$)
+    \item[Printer/copier stations] 1167 \kgcoe\ (\textcite{ecoinvent})
     \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
@@
     \item Deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region .
-    \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 kg CO$_2$e for usage,
-      97 kg CO$_2$e for manufacturing.
+    \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 \kgcoe\ for usage,
+      97 \kgcoe\ for manufacturing.
 \end{itemize}
@@
 
-Assuming a relatively clean grid at 200 g CO$_2$e / kWh (a low estimate), and that each server has embodied emissions of 4 t CO$_2$e (a high estimate),
+Assuming a relatively clean grid at 200 \gcoekwh (a low estimate), and that each server has embodied emissions of 4 t \coe\ (a high estimate),
 we obtain:
 \begin{itemize}
-    \item 100 t CO$_2$e / day operational emissions.
-    \item 260 t CO$_2$e training emissions.
-    \item 16 kt CO$_2$e total embodied emissions.
+    \item 100 t \coe\ / day operational emissions.
+    \item 260 t \coe\ training emissions.
+    \item 16 kt \coe\ total embodied emissions.
 \end{itemize}
 
-After one year of operation, operational emissions are approximately 36 kt CO$_2$e, whereas training and embodied emissions are about 16 kt CO$_2$e,
+After one year of operation, operational emissions are approximately 36 kt \coe, whereas training and embodied emissions are about 16 kt \coe,
 so operational emissions account for $\approx 70\%$ of the total. With less conservative assumptions (dirtier grid, lower embodied emissions),
@@
 
+\subsection{Model validation}
+
+Using Ecoinvent embodied emissions values where available, and only the numbers and types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
+Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe/device for embodied emissions, \kgcoe/device/day for operational emissions) provided by Ecoinvent.
+
+While this is admittedly just one data point, it provides some confidence that our assumptions are realistic and that the model can be used to extrapolate emissions and trends beyond the scope of the sustainability report.
+
 \subsection{Limitations and validation}
@@
 First, the model correctly estimates approximately $400$ kW of compute for central campus data centers.
-Second, for university-owned campus IT infrastructure, the model estimates $1.4$ kt CO$_2$e emissions per year, close to the $1.45$ kt reported in the 2024 university sustainability report \cite{uzh:sustainability:report}.
+Second, for university-owned campus IT infrastructure, the model estimates $1.4$ kt \coe\ emissions per year, close to the $1.45$ kt reported in the 2024 university sustainability report \cite{uzh:sustainability:report}.
 
@@
 \item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint.
-For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ t CO$_2$e annually.
+For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ t \coe\ annually.
 Usage-related emissions amount to about a tenth of that.
@@
 \item Embodied emissions are an important consideration for university-owned devices as well.
-If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ t CO$_2$e, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.
+If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ t \coe, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.
 
@@
 UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders.
-Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ t CO$_2$e annually, though this estimate requires further validation.
+Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ t \coe\ annually, though this estimate requires further validation.
 

### 2025-08-03 doc/paper.tex
Operational and embodied emissions of campus devices

@@
 \renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
-\newcommand{\assumption}[1]{{#1}}
 \newcommand{\eg}{{\em e.g.}}
@@
 
-On top of this, we assume that
+In addition to these numbers, which cover staff and facilities, we assume that 80\% of the student population of 28,000 people have a laptop (or equivalent device, such as a tablet), and that everyone at the university has mobile phone, resulting in a rough estimate of 32,000 laptops and 35,000 mobile phones.
+
+Traditional desktop computers appear to be infrequent on campus. By default we set their number to zero in our model, but we still provide parameters for them.
 
@@
 
+The Swiss electricity mix is substantially cleaner than the European or North American average, so generic manufacturer estimates of lifetime usage-related emissions are unsuitable.
+Unfortunately, in manufacturer literature we found surprisingly few sources of data on device power consumption, as opposed to estimated lifetime CO$_2$ emissions.
+Based on some manufacturer data sheets, primarily from Apple, Cisco, and Dell, we model power draw of different devices while under load using the following initial parameters:
+
+\begin{center}
+\begin{tabular}{|l|r|r|}
+\hline
+\textbf{Device} & \textbf{Mean (W)} & \textbf{$\sigma$ (W)} \\ \hline
+Laptop & 30 & 5 \\ \hline
+Monitor & 50 & 10 \\ \hline
+Desktop & 100 & 20 \\ \hline
+Server & 400 & 100 \\ \hline
+Mobile phone & 5 & 2 \\ \hline
+A/V equipment & 250 & 50 \\ \hline
+Printer & 1000 & 200 \\ \hline
+\end{tabular}
+\end{center}
+
+Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays, (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day.
+
+Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.
+
+We lack good data about the campus network, save for the existence of approximately 4,000 wifi access points.
+Based on this, we choose to model network emissions as a configurable percentage of device (local, desktop, server) emissions, by default 10\%.
+
 \subsubsection*{Embodied emissions}
 
-We obtain
+For several device types, especially end-user devices such as phones and laptops, embodied emissions
+comprise a majority of lifetime emissions.
+Different sources report widely different values for these emissions.
+Our approach has been to find several studies and industry references, and then estimate a mean reference value as well as likely minimum and maximum values.
+
+The following table summarizes results from the literature.
+
+\begin{center}
+\begin{tabular}{|l|r|l|}
+\hline
+& \textbf{Embod. emiss.} & \\
+\textbf{Device type} & \textbf{(\kgcoe)} & \textbf{Source} \\ \hline
+Laptop & 121 & \textcite{ecoinvent}  \\ \hline
+Laptop & 104 & \textcite{teehan2013} (low) \\ \hline
+Laptop & 338 & \textcite{teehan2013} (high) \\ \hline
+Laptop & 244 & \textcite{rarecoil} \\ \hline
+Laptop & 202 & \textcite{unctadder2024} \\ \hline
+Desktop & 238 & \textcite{ecoinvent} \\ \hline
+Desktop & 303 & \textcite{teehan2013} \\ \hline
+Desktop & 403 & \textcite{unctadder2024} \\ \hline
+Desktop & 289 & \textcite{dellpcf} \\ \hline
+Desktop & 277 & \textcite{boavizta} \\ \hline
+Server & 383 & \textcite{teehan2013} \\ \hline
+Server & 1252 & \textcite{davy2021} (Dell) \\ \hline
+Server & 1912 & \textcite{davy2021} (EC2) \\ \hline
+Server & 899 & \textcite{boavizta} (medium server)\\ \hline
+Phone & 68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
+Phone & 54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
+Phone & 50 & \textcite{unctadder2024} \\ \hline
+Phone & 50 & \textcite{lovehagen2023} \\ \hline
+\end{tabular}
+\label{tab:embodied_emissions}
+\end{center}
+
+For laptops, the numbers from Ecoinvent, a reputable proprietary Swiss life cycle inventory database used in the UZH sustainability report, appear to be on the low end.
+The most comprehensive publicly available data comes from \textcite{rarecoil}, which lists manufacturer information for over 90 popular devices.
+Mean supply chain emissions in this data set are 244 \kgcoe\ with a standard deviation of 128 kg.
+To be conservative, we split the difference between this number and the Ecoinvent values and model laptops embodied emissions at 180 \kgcoe\ / device.
+
+For desktops, most estimates are in the range of 250 - 400 \kgcoe.
+The most detailed data comes from Dell's Product Carbon Footprints (\cite{dellpcf}), which the company publishes for each of their products.
+A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg.
+We model desktop embodied emissions at 288 \kgcoe\ / device.
+
+In the case of server, \textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
+The former have mean supply chain emissions of 1252 \kgcoe\ with $\sigma = 330$ kg, the latter 1912 kg with $\sigma = 885$ kg.
+\textcite{boavizta} reports mean supply chain emissions of 899 \kgcoe\ for a medium-sized server.
+We model server supply chain emissions at 1252 \kgcoe\ / device.
+
+The range of values for mobile phones is much smaller. We model phone embodied emissions at 50 \kgcoe\ / device.
+
+We found less data for other types of devices. For now, we model them as follows:
+
+\begin{itemize}
+    \item Computer monitors: 344 \kgcoe\ (\cite{teehan2013}, \cite{dellpcf}).
+    \item Conference room displays: 753 \kgcoe\ (scaling monitor by $(40''/27'')^2$).
+We use this as a proxy for all audiovisual equipment (lecture room projectors, etc.).
+Note that the numbers are sufficiently low that errors in this estimate do not significantly affect the overall result.
+    \item Printer/copier stations: 1167 \kgcoe\ (\cite{ecoinvent}).
+    \item Network equipment: We did not find reliable lifecycle assessments for routers and switches,
+      and have little information about UZH network architecture. The vast majority ($80-95\%$) of
+      lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
+      For now, we model the embodied emissions of network equipment like we do its operational emissions, namely as a 10\% overhead on the number of devices (laptops, desktops, servers).
+\end{itemize}
 
@@
 in the literature about the energy consumption of each query.
-
-\subsection{Embodied emissions}
-
-For several device types, especially end-user devices such as phones and laptops, embodied emissions
-comprise a majority of lifetime emissions. However, the exact amount of these emissions varies
-widely between different sources. Our approach has been to find several studies and industry references,
-and then estimate means and standard deviations from those that seemed most comprehensive and reliable
-(for example, Dell's Product Carbon Footprint reports about its servers).
-
-\subsubsection{Laptops}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
-121 & \textcite{ecoinvent}  \\ \hline
-104 & \textcite{teehan2013} (low) \\ \hline
-338 & \textcite{teehan2013} (high) \\ \hline
-244 & \textcite{rarecoil} \\ \hline
-202 & \textcite{unctadder2024} \\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:laptops}
-\end{center}
-
-The Ecoinvent numbers appear to be outliers on the low end.
-The most comprehensive data comes from \textcite{rarecoil}, which lists manufacturer
-information for over 90 popular devices. Mean supply chain emissions are
-244 \kgcoe\ with a standard deviation of 128 kg.
-\assumption{We model laptop supply chain emissions using a truncated normal distribution with those characteristics.}
-
-\subsubsection{Desktops}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
-238 & \textcite{ecoinvent} \\ \hline
-303 & \textcite{teehan2013} \\ \hline
-403 & \textcite{unctadder2024} \\ \hline
-289 & \textcite{dellpcf} \\ \hline
-277 & \textcite{boavizta} \\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:desktops}
-\end{center}
-
-Most estimates are in the range of 250 - 400 \kgcoe. The most detailed data comes from
-Dell's Product Carbon Footprints (\cite{dellpcf}), which they publish for each of their products.
-A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg.
-\assumption{We model desktop supply chain emissions using these statistics.}
-
-\subsubsection{Servers}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
-383 & \textcite{teehan2013} \\ \hline
-1252 & \textcite{davy2021} (Dell) \\ \hline
-1912 & \textcite{davy2021} (EC2) \\ \hline
-899 & \textcite{boavizta} (medium server)\\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:servers}
-\end{center}
-
-\textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
-The former have mean supply chain emissions of 1252 \kgcoe\ with $\sigma = 330$ kg, the latter
-1912 kg with $\sigma = 885$ kg.
-\assumption{We model server supply chain emissions using the statistics corresponding to Dell servers.}
-
-\subsubsection{Smartphones}
-
-\begin{center}
-\begin{tabular}{|r|l|}
-\hline
-\textbf{Production \& Disposal} & \\
-\textbf{(\kgcoe)} & \textbf{Source} \\ \hline
-68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
-54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
-50 & \textcite{unctadder2024} \\ \hline
-50 & \textcite{lovehagen2023} \\ \hline
-\end{tabular}
-\label{tab:embodied_emissions:phones}
-\end{center}
-
-\assumption{We model phone supply chain emissions using mean 50 \kgcoe\ and $\sigma = 10$ kg.}
-
-\subsubsection{Other devices}
-
-We found less data for other types of devices. For now, we model them with averages based
-on a handful of sources.
-
-\begin{description}
-    \item[Computer monitors] 344 \kgcoe\ (\textcite{teehan2013}, \textcite{dellpcf})
-    \item[Conference room displays] 753 \kgcoe\ (scaling monitor by $(40''/27'')^2$)
-    \item[Printer/copier stations] 1167 \kgcoe\ (\textcite{ecoinvent})
-    \item[Network equipment] We did not find reliable lifecycle assessments for routers and switches,
-      and have no information about UZH network architecture. The vast majority ($80-95\%$) of
-      lifecycle emissions of network gear stem from usage (\textcite{cisco2024}, \textcite{jacob2023}).
-      For now, we model network equipment as a 5\% overhead on servers (1 router or switch for 20 servers).
-      (A typical data center rack contains 42 1U servers and one switch or router.)
-\end{description}
-
-\subsection{Operational emissions}
-
-For on-premise equipment, we would like to model the Swiss electricity mix (see below) and, ideally,
-daily variations in usage and electricity carbon intensity.
-Generic manufacturer estimates of lifetime usage-related emissions are therefore unsuitable.
-There are surprisingly few sources of data on device power consumption (as opposed to CO$_2$ emissions).
-Based on a smattering of data sheets, primarily from Apple, Cisco, and Dell, we model
-power draw of different devices while under load using the following initial parameters:
-
-\begin{center}
-\begin{tabular}{|l|r|r|}
-\hline
-\textbf{Device} & \textbf{Mean (W)} & \textbf{$\sigma$ (W)} \\ \hline
-Laptop & 30 & 5 \\ \hline
-Desktop & 100 & 20 \\ \hline
-Server & 400 & 100 \\ \hline
-Phone & 5 & 2 \\ \hline
-Monitor & 50 & 10 \\ \hline
-Conference display & 250 & 50 \\ \hline
-Printer/copier & 1000 & 200 \\ \hline
-\end{tabular}
-\end{center}
-
-We assume a distribution of duty cycles and power draw when idle for each device.
-In the future, we may simulate power draw at different times of day to integrate with
-hourly electricity mix data, but errors in those assumptions may swamp the relatively small
-hourly variations in Swiss electricity carbon intensity.
 

### 2025-08-03 doc/paper.tex
Many changes, 3rd party cloud, validation section in results

@@
+% boavizta mention immaturity of assessments and variability of estimates
+
 \documentclass[11pt]{article}
@@
 
+For servers, which are mostly located in on-premise data centers, we also attempt to take into account emissions from cooling.
+
 \subsubsection*{Cloud and supercomputing facilities}
@@
 UZH at present does not make significant use of commercial cloud compute, but the model allows simulating the shift of some fraction of the server load to cloud VMs.
+For this hypothetical cloud footprint, like for on-premise data centers, we attempt to model operational emissions, embodied emissions, and emissions from cooling.
 
-Due to lack of data, we do not attempt to model emissions from application-level cloud services such as Microsoft 365 or Google Workspace.
-This omission leads to undercounting total IT energy use but to overestimating the potential relative impact of AI.
+UZH faculties do use shared off-campus supercomputing facilities, primarily the Swiss National Supercomputing Center, CSCS~\cite{cscs}.
+We estimate the operational impact of these workloads, but, due to lack of data, we do not include embodied emissions or cooling.
+This omission undercounts total university IT energy use but overestimates the potential relative impact of AI.
 
-UZH faculties do make use of shared off-campus supercomputing facilities, primarily the Swiss National Supercomputing Center, CSCS~\cite{cscs}.
-We estimate the operational impact of these workloads, but, again due to lack of data, we do not include embodied emissions.
-Again, this undercounts total university IT energy use but overestimates the potential relative impact of AI.
+Due to lack of data, we do not model emissions from application-level cloud services such as Microsoft 365 or Google Workspace.
+Again, this undercounts total IT energy use but overestimate the relative impact of AI.
 
@@
 We estimate inference-time emissions of current large language models based on the literature.
-We also include training-time emissions and embodied emissions of language models (thouse these are harder to estimate and we err on the side of overestimating them).
+We also include training-time emissions, embodied emissions, and likely cooling overhead of language models (these are harder to measure and we err on the side of overestimating them).
 
@@
 \begin{center}
-  \begin{tabular}{|l|c|c|}
+  \begin{tabular}{|l|c|c|c|}
     \hline
-  \textbf{Device or service} & \textbf{Operation} & \textbf{Lifecycle} \\ \hline
-  Student laptops \& phones & yes & yes \\
-  University workstations, access points, etc. & yes & yes \\
-  Cloud servers & yes & yes \\
-  Cloud services & no & no \\
-  Supercomputing facilities & yes & no \\
-  LLMs & yes & yes \\ \hline
+    \textbf{Device or service} & \textbf{Operation} & \textbf{Lifecycle} & \textbf{Cooling} \\ \hline
+    Student laptops \& phones & yes & yes & n/a \\
+    University workstations, access points, etc. & yes & yes & n/a \\
+    University on-premise servers & yes & yes & yes \\
+    Cloud servers & yes & yes & yes \\
+    Cloud services & no & no & no \\
+    Supercomputing facilities & yes & no & no \\
+    LLMs & yes & yes & yes \\ \hline
   \end{tabular}
@@
 Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (\kgcoe).
+We then estimate the local grid's emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number.
+(In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution.
+However, we have insufficient data on the precise form of such a distribution, so we limit our analysis to the linear mode.)
 
-We then estimate the local grid's emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number.
-We obtain emissions intensity data from Electricity Maps~\cite{electricitymaps}.
+Note that for Switzerland we can compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps}.
+Emissions vary from a high of 64 \gcoekwh\ in the middle of the night to a low of 45 \gcoekwh\ in the middle of the day.
+This variation is sufficiently large that we have refined the model's device duty cycle to represent four six-hour periods over the course of the day.
+It allows us to specify, for example, that user devices are mostly used during the day whereas servers run around the clock.
 
-In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution.
-However, we have insufficient data on the precise form of such a distribution, so we limit our analysis to the linear mode.
+For servers (in on-premise data centers) we also account for cooling overhead.
+Based on limited UZH historical power data, we estimate that the on-premise power usage effectiveness (PUE), the ratio of total data center power (including cooling) to power used for computation, is approxmiately 2.
+This value, higher than the commercial average of 1.56~\cite{google:datacenter:efficiency}, is attributable to inefficient cooling.
 
@@
 
-We attempt to model a hypothetical shift of some workloads from servers in an on-premise data center to a commercial cloud. The parameters under consideration include:
-
-\begin{itemize}
-  \item On-premise power usage effectiveness (PUE), the ratio of total data center power (including cooling) to power used for computation.
-  \item Cloud data center PUE and grid emissions intensity (since the cloud data center may be located in a different region than the on-premise data center).
-  \item An average conversion ratio of on-premise servers to cloud servers.
-\end{itemize}
-
-We measure the impact of cloud servers in terms of annual operational emissions and amortized emissions due to manufacturing and disposal, in units of \kgcoe\ / year.
+We attempt to model a hypothetical shift of some workloads from servers in an on-premise data center to a commercial cloud.
+We model the shift as a fraction of servers that are moved to the cloud, the average conversion ratio of on-premise servers to cloud servers, and cloud data center PUE and electricity emissions intensity (since the cloud data center may be located in a different region than the on-premise data center).
 
@@
 
-\subsection{Parameters and assumptions}
-
-The model's outputs are affected by several factors, and where we were unable to find good data we made assumptions that we considered reasonable.
-All of these parameters and assumptions are parameterizable by users of the model.
-We list these sources of uncertainty below.
-
-\begin{description}
-  \item[Cloud data center PUE] As of 2025, industry average power usage effectiveness is 1.56, with Google averaging 1.09 across its fleet~\cite{google:datacenter:efficiency} and Microsoft ranging from 1.12 to 1.30~\cite{microsoft:datacenter:efficiency}.
-We choose a PUE of 1.2 as representative of a state-of-the-art hyperscaler data center.
-  \item[On-premise data center PUE] Based on limited UZH historical power data, we assume an on-premise PUE of 2.0, attributable to inefficient cooling.
-  \item[Device duty cycle] We make unsupported but plausible assumptions about the number of days that different devices are used (\eg, end-user devices are not used every day) and their duty cycle throughut the day.
-  \item[Electricity emissions intensity for on-premise data centers] For Switzerland, we can compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps}.
-Emissions vary from a high of 64 \gcoekwh\ in the middle of the night to a low of 45 \gcoekwh\ in the middle of the day.
-This variation is sufficiently large that we refine the model's device duty cycle to four six-hour periods over the course of the day.
-It allows us to specify, for example, that user-devices are mostly used during the day whereas servers run around the clock.
-  \item[Electricity emissions intensity for cloud data centers and LLMs] These facilities are unlikely to be located in Switzerland, and Switzerland's grid is cleaner than average.
-We use the European Union's average emissions intensity of electricity generation in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
-  \item[LLM usage] We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).
-\end{description}
-
 \section{Model details}
@@
 The previous section describes our overall modeling approach, which we believe is broadly applicable to university and other large campus settings.
-This section presents the reasoning behind specific default values we employ in the UZH evaluation.
+This section presents the reasoning behind specific values we employ in the UZH evaluation.
 
@@
 
-Traditional desktop computers appear to be infrequent on campus. By default we set their number to zero in our model, but we still provide parameters for them.
+Traditional desktop computers appear to be infrequent on campus. By default we set their number to zero in our model, but we still provide the framework to model them if desired.
 
@@
 The Swiss electricity mix is substantially cleaner than the European or North American average, so generic manufacturer estimates of lifetime usage-related emissions are unsuitable.
-Unfortunately, in manufacturer literature we found surprisingly few sources of data on device power consumption, as opposed to estimated lifetime CO$_2$ emissions.
+Unfortunately, in manufacturer documentation we found surprisingly few sources of data on device power consumption, as opposed to estimated lifetime CO$_2$ emissions.
 Based on some manufacturer data sheets, primarily from Apple, Cisco, and Dell, we model power draw of different devices while under load using the following initial parameters:
@@
 Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays, (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day.
-
 Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.
@@
     Desktop & 289 & \textcite{dellpcf} \\ \hline
-Desktop & 277 & \textcite{boavizta} \\ \hline
+    Desktop & 277 & \textcite{boavizta:api} \\ \hline
     Server & 383 & \textcite{teehan2013} \\ \hline
@@
     Server & 1912 & \textcite{davy2021} (EC2) \\ \hline
-Server & 899 & \textcite{boavizta} (medium server)\\ \hline
+    Server & 899 & \textcite{boavizta:api} (medium server)\\ \hline
     Phone & 68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
@@
 
-In the case of server, \textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances.
+In the case of servers, \textcite{davy2021} reports manufacturer data for many Dell servers and estimates for different types of EC2 instances.
 The former have mean supply chain emissions of 1252 \kgcoe\ with $\sigma = 330$ kg, the latter 1912 kg with $\sigma = 885$ kg.
-\textcite{boavizta} reports mean supply chain emissions of 899 \kgcoe\ for a medium-sized server.
+\textcite{boavizta:api} reports mean supply chain emissions of 899 \kgcoe\ for a medium-sized server.
 We model server supply chain emissions at 1252 \kgcoe\ / device.
 
-The range of values for mobile phones is much smaller. We model phone embodied emissions at 50 \kgcoe\ / device.
+The variability of estimates for mobile phones is much smaller. We model phone embodied emissions at 50 \kgcoe\ / device.
 
@@
   \item Printer/copier stations: 1167 \kgcoe\ (\cite{ecoinvent}).
-    \item Network equipment: We did not find reliable lifecycle assessments for routers and switches,
-      and have little information about UZH network architecture. The vast majority ($80-95\%$) of
-      lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
+  \item Network equipment: We did not find reliable lifecycle assessments for routers and switches, and have little information about UZH network architecture.
+The vast majority ($80-95\%$) of lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
 For now, we model the embodied emissions of network equipment like we do its operational emissions, namely as a 10\% overhead on the number of devices (laptops, desktops, servers).
@@
 
-\subsection{LLMs}
+The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud infrastructure.
+The best resource we have found is the Datavizta API by the Boavizta working group~\cite{boavizta:api}.
 
-We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
-to cover the rest of the university population.
+Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as the replacement of one campus server with a configurable number (by default, one) of Azure {\tt standard\_d48s\_v3} instances (Intel Xeon, 48 cores, 192 GB RAM, 2TB SSD)~\cite{msftvms}.
+Based on data from Boavizta, such a machine has embodied emissions of 667 \kgcoe\ and mean electricity power usage of 255 W.
+We assume deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region, which makes sense for a Swiss university and means that the electricity mix is the same as for on-premise facilities.
+We estimate a power usage effectiveness of 1.15 for this facility.
+(As of 2025, average PUE for cloud computing is 1.56, with Google averaging 1.09 across its fleet~\cite{google:datacenter:efficiency} and Microsoft ranging from 1.14 to 1.19 in Europe~\cite{microsoft:datacenter:efficiency}.)
 
-For each device instance, we pick embodied emissions, power draw, and duty cycles either
-deterministically or by sampling from the appropriate distribution.
+Given all these parameters, we can then compute cloud server operational emissions, embodied emissions, and cooling overhead as we do for on-premise servers.
 
-Combining power draw and duty cycle estimates with ground-truth data about the Swiss electricity mix,
-we estimate annual operational emissions and add them to the embodied emissions scaled by the device
-type's expected lifetime.
+\subsection{Large language models}
 
-For on-campus data centers, for which presumably there are electricity bills or other measures of
-overall energy use, we would like to use that data to evaluate our model.
-
-A possible future refinement will be to model the duty cycle as a function of time of day
-and incorporate finer-grained electricity mix data.
-
-\subsection{Cloud emissions}
-
-We model cloud emissions based on the assumptions and data described in
-Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
-more accurate data.
-
-\subsection{LLM emissions}
+We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).
 
@@
 
-\subsection{Cloud data centers}
-\label{sec:cloud_data_centers}
+Energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring}.
 
-The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
-infrastructure. The best resource we have been able to find is the Datavizta API (\cite{boavizta}).
-
-Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as follows:
-\begin{itemize}
-    \item One-for-one replacement of one campus server with one Azure D8S\_v3 instance (2x Intel
-      Xeon, 8 cores, 32 GB RAM, 2TB SSD) (\cite{msftvms}).
-    \item Deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region .
-    \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 \kgcoe\ for usage,
-      97 \kgcoe\ for manufacturing.
-\end{itemize}
-
-\subsection{Large language models}
-
-Energy consumption of LLMs is being studied extensively (\cite{budennyy2022eco2ai},
-\cite{castano2023exploring},
-\cite{devries2023growing},
-\cite{gowda2024watt},
-\cite{harding2024watts},
-\cite{heguerte2023estimate},
-\cite{luccioni2022estimating},
-\cite{luccioni2023counting},
-\cite{patterson2021carbon},
-\cite{rodriguez2024evaluating},
-\cite{tripp2024measuring}).
-
-The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request,
-or approximately 10x the energy of a traditional search engine query. This number appears to be decreasing
-rapidly as model hardware and software improve, even though overall energy use is increasing due to
-increased utilization. Nevertheless, for the model we assume that queries consume 3 Wh each.
+The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request, or approximately 10x the energy of a traditional search engine query.
+This number appears to be decreasing rapidly as model hardware and software improve, even though overall energy use is increasing due to increased utilization.
+Nevertheless, for the model we assume that queries consume 3 Wh each.
 We assume that on average every user will issue 15 queries per day.
 
-LLMs are distributed and the location that serves any particular request is unknown, so we use the
-average European electricity mix to compute carbon intensity.
+LLMs are distributed and the location that serves any particular request is unknown, so we use the average European electricity mix to compute carbon intensity.
 
-The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
-However, \textcite{devries2023growing} provides the following information about ChatGPT:
+The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely. However, \textcite{devries2023growing} provides the following information about ChatGPT:
 \begin{itemize}
@@
 
-Assuming a relatively clean grid at 200 \gcoekwh (a low estimate), and that each server has embodied emissions of 4 t \coe\ (a high estimate),
-we obtain:
+Assuming a relatively clean grid at 200 \gcoekwh (a low estimate), and that each server has embodied emissions of 4 t \coe\ (a high estimate), we obtain:
 \begin{itemize}
@@
 
-After one year of operation, operational emissions are approximately 36 kt \coe, whereas training and embodied emissions are about 16 kt \coe,
-so operational emissions account for $\approx 70\%$ of the total. With less conservative assumptions (dirtier grid, lower embodied emissions),
-the proportion of operational emissions increases further.
+After one year of operation, operational emissions are approximately 36 kt \coe, whereas training and embodied emissions are about 16 kt \coe, so operational emissions account for $\approx 70\%$ of the total.
+With less conservative assumptions (dirtier grid, lower embodied emissions), the proportion of operational emissions increases further.
+
+Electricity emissions intensity for cloud data centers and LLMs] These facilities are unlikely to be located in Switzerland, and Switzerland's grid is cleaner than average.
+We use the European Union's average emissions intensity of electricity generation in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
 
@@
 
-The Swiss Federal Office for the Environment published an extensive report (\cite{krebs2018umweltbilanz})
-on the environmental impact of electricity generation in Switzerland in 2018.
+The Swiss Federal Office for the Environment published an extensive report (\cite{krebs2018umweltbilanz}) on the environmental impact of electricity generation in Switzerland in 2018.
 
-However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
-we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
-at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
-hour block over the course of the year. This gives us twelve data points that describe average
-carbon intensity variation over the course of the day.
-
-\subsection{Model validation}
-
-Using Ecoinvent embodied emissions values where available, and only the numbers and types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
-Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe/device for embodied emissions, \kgcoe/device/day for operational emissions) provided by Ecoinvent.
-
-While this is admittedly just one data point, it provides some confidence that our assumptions are realistic and that the model can be used to extrapolate emissions and trends beyond the scope of the sustainability report.
+However, for the purposes of the model, we use data from \textcite{electricitymaps}.
+Specifically, we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023 at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two hour block over the course of the year.
+This gives us twelve data points that describe average carbon intensity variation over the course of the day.
 
@@
 \label{sec:results}
+
+\subsection{Model validation}
+
+Using Ecoinvent embodied emissions values where available, and only the numbers and types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
+Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe/device for embodied emissions, \kgcoe/device/day for operational emissions) provided by Ecoinvent.
+
+While this is admittedly just one data point, it provides some confidence that our assumptions are realistic and that the model can be used to extrapolate emissions and trends beyond the scope of the sustainability report.
 

### 2025-08-03 doc/paper.tex
Checkpoint: LLMs, results structure, conclusion bits

@@
 
+For supercomputing at CSCS, unfortunately, our only available data is power utilisation at peak load, 131 kW in 2024.
+We assume a 50\% average load and use the Swiss emissions intensity numbers to compute annual operational emissions.
+
 \subsection{Large language models}
 
-We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).
-
-We model LLM emissions by sampling from the per-capita daily query distribution and computing
-the carbon intensity of each query based on the average European electricity mix. While there
-are many uncertainties about the level of LLM adoption, there appears to be reasonable consensus
-in the literature about the energy consumption of each query.
-
-Energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring}.
+The energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring}.
 
@@
 Nevertheless, for the model we assume that queries consume 3 Wh each.
-We assume that on average every user will issue 15 queries per day.
-
-LLMs are distributed and the location that serves any particular request is unknown, so we use the average European electricity mix to compute carbon intensity.
 
@@
 
-Assuming a relatively clean grid at 200 \gcoekwh (a low estimate), and that each server has embodied emissions of 4 t \coe\ (a high estimate), we obtain:
+The facilities hosting LLMs are distributed and the location that serves any particular request is unknown, so to map energy use to emissions we use the average emissions intensity of electricity generation in the EU in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
+
+Assuming this emissions intensity, and that each server has embodied emissions of 4 t \coe\ (a high estimate), we obtain:
 \begin{itemize}
@@
 
-After one year of operation, operational emissions are approximately 36 kt \coe, whereas training and embodied emissions are about 16 kt \coe, so operational emissions account for $\approx 70\%$ of the total.
+After one year of operation, operational emissions are approximately 36 kt \coe, whereas training and embodied emissions are about 16 kt \coe, so operational emissions account for $\sim 70\%$ of the total.
 With less conservative assumptions (dirtier grid, lower embodied emissions), the proportion of operational emissions increases further.
 
-Electricity emissions intensity for cloud data centers and LLMs] These facilities are unlikely to be located in Switzerland, and Switzerland's grid is cleaner than average.
-We use the European Union's average emissions intensity of electricity generation in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
+As a result, when modeling overall LLM emissions, we multiply estimated operational emissions by 1.4 to account for training and embodied emissions.
+(We acknowledge that this is a rough estimate based on just one LLM.)
 
-As a result, when modeling overall LLM emissions, we multiply estimated operational emissions by 1.4 to account for training and embodied emissions.
-We acknowledge that this is, at best, a rough estimate based on just one LLM.
+We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day). We assume that on average every user will issue 15 queries per day.
 
@@
 
-\subsection{Limitations and validation}
+\subsection{Model limitations}
 
-The biggest limitation is the modest availability of some ground truth data about UZH IT infrastructure.
-Specifically, we have only estimates of campus data center energy use, and limited information abut UZH research workloads (whether AI or not).
+Our biggest limitation is the relative lack of ground truth data about UZH IT infrastructure.
+We have obtained only estimates of campus data center energy use, and little information abut UZH research workloads (whether AI-related or not).
 We assume that many research AI workloads will be relatively similar in intensity to existing scientific computation workloads.
-We also have limited information (e.g., peak power consumption) about UZH use of external high-performance compute resources such as Alps/CSCS.
+We also lack information about UZH use of external high-performance compute resources at CSCS.
 
-Additionally, cloud vendors and AI companies are secretive about their infrastructure. We have relied on public datasets and research literature to estimate cloud and AI emissions.
+Additionally, our work relies on assumptions and anecdotal evidents about device ownership and use among the university population, use of generative AI tools, etc.
 
-Despite these limitations, our bottom-up model aligns closely with two UZH data points.
-First, the model correctly estimates approximately $400$ kW of compute for central campus data centers.
-Second, for university-owned campus IT infrastructure, the model estimates $1.4$ kt \coe\ emissions per year, close to the $1.45$ kt reported in the 2024 university sustainability report \cite{uzh:sustainability:report}.
+Moreover, cloud vendors and AI companies are secretive about their infrastructure, and we have had to rely on public datasets and research literature to estimate cloud and AI emissions.
+
+However, all these points of uncertainty are exposed as user-configurable parameters in the model, allowing easy sensitivity analysis and exploration of alternate scenarios.
 
@@
 
-\subsection{Model validation}
+\subsection{Initial model validation}
 
 Using Ecoinvent embodied emissions values where available, and only the numbers and types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
-Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe/device for embodied emissions, \kgcoe/device/day for operational emissions) provided by Ecoinvent.
+Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.
 
-While this is admittedly just one data point, it provides some confidence that our assumptions are realistic and that the model can be used to extrapolate emissions and trends beyond the scope of the sustainability report.
+While this is admittedly just one data point, and the extreme proximity of the two values is likely a coincidence, it provides some confidence that our assumptions are reasonable and that the model can be used to make directional decisions and extrapolate trends beyond the scope of the sustainability report.
+
+\subsection{Key results}
+
+\subsubsection*{Impact of embodied emissions}
+
+\subsubsection*{Impact of cloud computing}
+
+\subsubsection*{Impact of LLMs}
+
+\subsection{Sensitivity analysis}
+
+\section{Recommendations}
+\label{sec:recommendations}
+
+Our recommendations for university policy makers can be summarized in three main points.
+
+First, do not panic about LLMs. Their emissions comprise a small fraction of total IT emissions, they will get more efficient, and attempting to curb their use is likely to be a losing battle.
+
+Invest in infrastructure for the long term.
+Run awareness campaigns and adapt IT policies so that personal devices and university IT systems are used for as long as possible.
+Amortize manufacturing emissions over the longest period possible.
+
+Improve the cooling efficiency of university data centers, or engage cloud vendors and academic compute centers (such as CSCS) to validate the potential benefits of a migration to cloud.
+
+\section{Related work}
+\label{sec:related:work}
+
+\section{Conclusion}
 
@@
 
-\section{Recommendations}
-\label{sec:recommendations}
-
-Do not panic about LLMs. Their emissions comprise a small fraction of total IT emissions, they will get more efficient, and attempting to curb their use is likely to be a losing battle.
-Invest in infrastructure for the long term.
-Run awareness campaigns and adapt IT policies so that personal devices and university IT systems are used for as long as possible.
-Amortize manufacturing emissions over the longest period possible.
-Improve the cooling efficiency of university data centers, or engage cloud vendors and academic compute centers (such as CSCS) to validate the potential benefits of a migration to cloud.
-
-\section{Related work}
-\label{sec:related:work}
-
-\section{Conclusion}
-
 \printbibliography

### 2025-08-04 doc/paper.tex
Some value tweaks/roundings

@@
 A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg.
-We model desktop embodied emissions at 288 \kgcoe\ / device.
+We model desktop embodied emissions at 290 \kgcoe\ / device.
 
@@
 \textcite{boavizta:api} reports mean supply chain emissions of 899 \kgcoe\ for a medium-sized server.
-We model server supply chain emissions at 1252 \kgcoe\ / device.
+We model server supply chain emissions at 1100 \kgcoe\ / device.
 
@@
 \begin{itemize}
-  \item Computer monitors: 344 \kgcoe\ (\cite{teehan2013}, \cite{dellpcf}).
-  \item Conference room displays: 753 \kgcoe\ (scaling monitor by $(40''/27'')^2$).
+  \item Computer monitors: 340 \kgcoe\ (\cite{teehan2013}, \cite{dellpcf}).
+  \item Conference room displays: 700 \kgcoe\ (scaling monitor by $(40''/27'')^2$).
 We use this as a proxy for all audiovisual equipment (lecture room projectors, etc.).
 Note that the numbers are sufficiently low that errors in this estimate do not significantly affect the overall result.
-  \item Printer/copier stations: 1167 \kgcoe\ (\cite{ecoinvent}).
+  \item Printer/copier stations: 1100 \kgcoe\ (\cite{ecoinvent}).
   \item Network equipment: We did not find reliable lifecycle assessments for routers and switches, and have little information about UZH network architecture.
@@
 Based on data from Boavizta, such a machine has embodied emissions of 667 \kgcoe\ and mean electricity power usage of 255 W.
+Given the uncertainty, we round up to 700 \kgcoe\ and 255 W.
 We assume deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region, which makes sense for a Swiss university and means that the electricity mix is the same as for on-premise facilities.

### 2025-08-04 doc/paper.tex
LLM additions

@@
 For servers, which are mostly located in on-premise data centers, we also attempt to take into account emissions from cooling.
+(We do not model the manufacturing emissions of the cooling equipment itself.)
 
@@
 
-The energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring}.
+The energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,epoch2024optimallyallocatingcomputebetweeninferenceandtraining,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring,epoch2025howmuchenergydoeschatgptuse}.
 
-The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request, or approximately 10x the energy of a traditional search engine query.
-This number appears to be decreasing rapidly as model hardware and software improve, even though overall energy use is increasing due to increased utilization.
-Nevertheless, for the model we assume that queries consume 3 Wh each.
+A widely cited 2023 study~\cite{devries2023growing} estimated the average energy consumption of a large language model like ChatGPT at approximately 3 Wh / request, about ten times the energy of a standard Google search query.
+In 2025, \textcite{epoch2025howmuchenergydoeschatgptuse} makes a convincing case that the energy use for the average GPT-4o query is just 0.3 Wh, ten times smaller than the 2023 estimate, while admitting that for unusually long inputs (10K--100K tokens) or for more ambitious applications such as OpenAI's Deep Research~\cite{oai:deepresearch} the energy requirement might rise into the several tens of watt-hours.
+
+Improvements in hardware technology and an increased focus on software optimization by AI companies as the adoption of AI grows means that, for a given level or quality of model, these costs are likely to continue to drop rapidly.
+Nonetheless, to be conservative, in the model we continue to assume that queries consume 3 Wh each on average.
 

### 2025-08-04 doc/paper.tex
Finished LLM section (and all model details)

@@
 Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (\kgcoe).
-We then estimate the local grid's emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number.
-(In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution.
-However, we have insufficient data on the precise form of such a distribution, so we limit our analysis to the linear mode.)
+We then estimate the local grid's carbon emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number\footnote{The model can also operate stochastically, instantiating devices from a distribution, but we lack good data on the  form of such a distribution, so we limit our analysis to this simple linear mode.}.
 
-Note that for Switzerland we can compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps}.
+The Swiss Federal Office for the Environment has an extensive report on the environmental impact of electricity generation in Switzerland~\cite{krebs2018umweltbilanz}, but we are fortunate to also have hourly emissions intensity data from \textcite{electricitymaps}.
+We use the hourly intensity data for {\em consumption} (not production) to compute emissions intensity for every hour of the day averaged over all days in 2024.
 Emissions vary from a high of 64 \gcoekwh\ in the middle of the night to a low of 45 \gcoekwh\ in the middle of the day.
-This variation is sufficiently large that we have refined the model's device duty cycle to represent four six-hour periods over the course of the day.
+This variation is sufficiently large that we refine the model's device duty cycle to represent four six-hour periods over the course of the day.
 It allows us to specify, for example, that user devices are mostly used during the day whereas servers run around the clock.
@@
 A widely cited 2023 study~\cite{devries2023growing} estimated the average energy consumption of a large language model like ChatGPT at approximately 3 Wh / request, about ten times the energy of a standard Google search query.
-In 2025, \textcite{epoch2025howmuchenergydoeschatgptuse} makes a convincing case that the energy use for the average GPT-4o query is just 0.3 Wh, ten times smaller than the 2023 estimate, while admitting that for unusually long inputs (10K--100K tokens) or for more ambitious applications such as OpenAI's Deep Research~\cite{oai:deepresearch} the energy requirement might rise into the several tens of watt-hours.
+In 2025, \textcite{epoch2025howmuchenergydoeschatgptuse} makes a convincing case that the energy use for the average GPT-4o query is now just 0.3 Wh, ten times smaller than the 2023 estimate, while admitting that for unusually long inputs (10K--100K tokens) or more ambitious applications such as OpenAI's Deep Research~\cite{oai:deepresearch} the energy requirement might rise into the tens of watt-hours.
 
-Improvements in hardware technology and an increased focus on software optimization by AI companies as the adoption of AI grows means that, for a given level or quality of model, these costs are likely to continue to drop rapidly.
-Nonetheless, to be conservative, in the model we continue to assume that queries consume 3 Wh each on average.
+Improvements in hardware technology and an increased focus on software optimization by AI companies as adoption of AI grows mean that, for a given level or quality of model, inference costs are likely to continue to drop rapidly.
+To be conservative, in the model we assume that queries consume 1 Wh each on average.
 
-The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely. However, \textcite{devries2023growing} provides the following information about ChatGPT:
+By contrast, due to the increasing size and complexity of state-of-the-art models, the power required to train such models has been doubling annually~\cite{epoch2024powerusagetrend}.
+The literature offers little insight into the relative electricity consumption of training and inference~\cite{verdecchia2023systematic}, but Google~\cite{patterson:footprint:2022} reports that in 2019--2021 inference accounted for about 60\% of AI energy consumption, implying a ratio of training energy to inference energy was 2:3.
+
+Another way to estimate training overhead is to consider that training Llama 3.1, a model comparable to GPT-4o, required approximately 25 MW of power for three months, or 54 GWh.
+Assuming ChatGPT's recently reported 2.5 billion queries per day~\cite{techcrunch:chatgpt} and a conservative (for this analysis) 0.3 Wh per query, ChatGPT consumes 750 MWh / day.
+If the model is used for 9 months before being replaced, inference energy amounts to 203 GWh, so the ratio of training to inference is 54:204 $\approx$ 1:4.
+(Assuming 1 Wh per query, inference energy is 675 GWh, and the ratio is 1:13, making training overhead negligible.) To be conservative, in the model we assume a ratio of 1:4.
+
+We must also examine embodied emissions.
+\textcite{epoch2025howmuchenergydoeschatgptuse} estimates 1 second of NVIDIA H100 GPU time per query.
+NVIDIA reports~\cite{nvidia:h100} a cradle-to-gate carbon footprint of 1,312 \kgcoe\ for one H100.
+GPUs are typically hosted in sets of 4 or 8 in servers such as the NVIDIA DGX~\cite{nvidia:dgx}.
+Assuming a high-end server footprint of 2,000 \kgcoe\ for such a device and dividing by 8 GPUs, we obtain 250 \kgcoe\ server overhead per GPU.
+If this platform is used for three years at 70\% efficiency, we straightforwardly obtain embodied emissions of 0.01 \gcoe\ per query.
+
+% 3 * 365 * 24 * 3600 / .7 = 135154285 queries
+% (1312+250)*1000 = 1562000 grams
+% 1562000/135154285 grams / query
+
+Finally, there is the question of electricity mix.
+The facilities hosting LLMs are distributed and the location that serves a particular request is unknown.
+To map energy use to emissions we use the average emissions intensity of electricity generation in the EU in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
+
+In summary, we make the following default assumptions:
+
 \begin{itemize}
-  \item Operational power consumption of 500 MWh / day.
-  \item Approximately 1300 MWh consumed in training.
-  \item Approximately 4000 servers.
+  \item LLM energy consumption of 1 Wh per query.
+  \item Training overhead of 25\%. In other words, inference plus training amounts to 1.25 Wh per query.
+  \item Carbon emission intensity of 207 \gcoekwh.
+  \item Embodied emissions of 0.01 \gcoe\ / query.
 \end{itemize}
 
-The facilities hosting LLMs are distributed and the location that serves any particular request is unknown, so to map energy use to emissions we use the average emissions intensity of electricity generation in the EU in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
-
-Assuming this emissions intensity, and that each server has embodied emissions of 4 t \coe\ (a high estimate), we obtain:
-\begin{itemize}
-  \item 100 t \coe\ / day operational emissions.
-  \item 260 t \coe\ training emissions.
-  \item 16 kt \coe\ total embodied emissions.
-\end{itemize}
-
-After one year of operation, operational emissions are approximately 36 kt \coe, whereas training and embodied emissions are about 16 kt \coe, so operational emissions account for $\sim 70\%$ of the total.
-With less conservative assumptions (dirtier grid, lower embodied emissions), the proportion of operational emissions increases further.
-
-As a result, when modeling overall LLM emissions, we multiply estimated operational emissions by 1.4 to account for training and embodied emissions.
-(We acknowledge that this is a rough estimate based on just one LLM.)
-
-We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day). We assume that on average every user will issue 15 queries per day.
-
-\subsection{Electricity carbon intensity}
-
-The Swiss Federal Office for the Environment published an extensive report (\cite{krebs2018umweltbilanz}) on the environmental impact of electricity generation in Switzerland in 2018.
-
-However, for the purposes of the model, we use data from \textcite{electricitymaps}.
-Specifically, we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023 at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two hour block over the course of the year.
-This gives us twelve data points that describe average carbon intensity variation over the course of the day.
+Of course, we must also make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).
+By default we assume that the average every user will issue 15 queries per day.
 
@@
 
-Additionally, our work relies on assumptions and anecdotal evidents about device ownership and use among the university population, use of generative AI tools, etc.
+Additionally, our work relies on assumptions and anecdotal evidence about device ownership and use among the university population, use of generative AI tools, etc.
 

### 2025-08-04 doc/paper.tex
Start of results

@@
 %\usepackage{courier} % Courier font for monospace
+\usepackage{graphicx}
 \usepackage{hyperref}
@@
 
-Using Ecoinvent embodied emissions values where available, and only the numbers and types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
+Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
 Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.
 
-While this is admittedly just one data point, and the extreme proximity of the two values is likely a coincidence, it provides some confidence that our assumptions are reasonable and that the model can be used to make directional decisions and extrapolate trends beyond the scope of the sustainability report.
+While this is admittedly just one data point, and the extreme proximity of the two values is likely a coincidence, it provides some confidence that our assumptions are reasonable and that the model can be used to make directional decisions and extrapolate trends beyond the immediate scope of the sustainability report.
 
@@
 
+We now discuss the main takeaways from our model when it is configured with the parameter values described above.
+In Section~\ref{sec:sensitivity:analysis} we explore how these results change under different assumptions.
+
 \subsubsection*{Impact of embodied emissions}
+
+Maybe the most striking result is that, when we consider the entire population of campus devices, and not only those owned and managed by the university, overall emissions are absolutely dominated by the manufacturing and disposal of end-user devices.
+See Figure~\ref{fig:1}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.8\textwidth]{fig1.png}
+\caption{Baseline results}
+\label{fig:1}
+\end{figure}
 
@@
 \subsection{Sensitivity analysis}
+\label{sec:sensitivity:analysis}
 

### 2025-08-04 doc/paper.tex
Results checkpoint

@@
 \newcommand{\gcoe}{g \coe}
+\newcommand{\tcoe}{t \coe}
 \newcommand{\kgcoe}{k\gcoe}
@@
 Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}.
-At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt \coe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
+At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 M\tcoe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
 The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).
@@
 
-Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
+Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 \tcoe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
 Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.
@@
   \centering
-\includegraphics[width=0.8\textwidth]{fig1.png}
+  \includegraphics[width=0.8\textwidth]{fig-baseline.png}
   \caption{Baseline results}
@@
 
+Many studies~\cite{wb:itu:ict,umich:factsheet} agree that a large fraction of computer and mobile phone energy demand stems from the manufacturing phase.
+This is due in part to the fact that such end-user devices are not used continuously.
+Our estimate of embodied emissions comprising approximately 85\% of total laptop emissions is near the top of the range in the literature, and it is consistent with the fact that the carbon intensity of Swiss electricity is about one quarter the EU average~\cite{electricitymaps}.
+
 \subsubsection*{Impact of cloud computing}
+
+Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170 \tcoe\ of cooling overhead, and replaces 415 \tcoe\ of total server emissions with 380 \tcoe\ of cloud emissions.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{fig-cloud.png}
+  \caption{Migration to cloud}
+  \label{fig:1}
+\end{figure}
+
+For relatively small and inefficiently cooled facilities such as those of UZH, migration to the cloud certainly improves power usage effectiveness.
+Whether the migration is a net win, however, depends on details that would require an in-depth engagement with the cloud provider.
+The results above assume a one-for-one swap of on-premise servers with cloud ones, as well as the same electricity emissions intensity for the two facilities.
+If the cloud footprint needed to be bigger, or if the data center were located in a region with dirtier electricity, those factors might outweight the improved cooling efficiency.
+
+There is also a financial argument to consider: if migration to the cloud is more expensive than continued on-premise operation, the savings from not migrating might be invested in more cost effective emissions improvements (including ones not related to IT).
 
@@
   \item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint.
-For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ t \coe\ annually.
+For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ \tcoe\ annually.
 Usage-related emissions amount to about a tenth of that.
@@
   \item Embodied emissions are an important consideration for university-owned devices as well.
-If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ t \coe, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.
+If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ \tcoe, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.
 
@@
 UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders.
-Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ t \coe\ annually, though this estimate requires further validation.
+Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ \tcoe\ annually, though this estimate requires further validation.
 

### 2025-08-04 doc/paper.tex
Finished results draft

@@
 For desktops, most estimates are in the range of 250 - 400 \kgcoe.
-The most detailed data comes from Dell's Product Carbon Footprints (\cite{dellpcf}), which the company publishes for each of their products.
+The most detailed data comes from Dell's Product Carbon Footprints~\cite{dellpcf}, which the company publishes for each of their products.
 A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg.
@@
 \begin{itemize}
-  \item Computer monitors: 340 \kgcoe\ (\cite{teehan2013}, \cite{dellpcf}).
+  \item Computer monitors: 340 \kgcoe~\cite{dellpcf,lovehagen2023,teehan2013}.
   \item Conference room displays: 700 \kgcoe\ (scaling monitor by $(40''/27'')^2$).
@@
 Note that the numbers are sufficiently low that errors in this estimate do not significantly affect the overall result.
-  \item Printer/copier stations: 1100 \kgcoe\ (\cite{ecoinvent}).
+  \item Printer/copier stations: 1100 \kgcoe~\cite{ecoinvent}.
   \item Network equipment: We did not find reliable lifecycle assessments for routers and switches, and have little information about UZH network architecture.
-The vast majority ($80-95\%$) of lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
+The vast majority ($80-95\%$) of lifecycle emissions of network gear stem from usage~\cite{cisco2024,jacob2023}.
 For now, we model the embodied emissions of network equipment like we do its operational emissions, namely as a 10\% overhead on the number of devices (laptops, desktops, servers).
@@
 Maybe the most striking result is that, when we consider the entire population of campus devices, and not only those owned and managed by the university, overall emissions are absolutely dominated by the manufacturing and disposal of end-user devices.
-See Figure~\ref{fig:1}.
+See Figure~\ref{fig:baseline}.
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{fig-baseline.png}
+  \includegraphics[width=0.7\textwidth]{fig-baseline.png}
   \caption{Baseline results}
-  \label{fig:1}
+  \label{fig:baseline}
 \end{figure}
@@
 Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170 \tcoe\ of cooling overhead, and replaces 415 \tcoe\ of total server emissions with 380 \tcoe\ of cloud emissions.
+See Figure~\ref{fig:cloud}.
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{fig-cloud.png}
+  \includegraphics[width=0.7\textwidth]{fig-cloud.png}
   \caption{Migration to cloud}
-  \label{fig:1}
+  \label{fig:cloud}
 \end{figure}
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\textwidth]{fig-llm-growth.png}
+  \caption{LLM growth projection}
+  \label{fig:llm:growth}
+\end{figure}
+
+AI adoption is growing rapidly, driving demand for data centers and potentially stressing electrical infrastructure in coming years~\cite{lbl:dc:2024,iea:ai:energy}.
+However, for an institution of roughly constant size such as UZH, and where IT infrastructure is already well developed and heavily used, emissions from generative AI most likely represent a small addition to overall emissions.
+In our model, using the default parameters, they comprise just 1\% of total emissions.
+
+Moreover, under even aggressive growth projections (50\% YoY usage growth, 50\% YoY training overhead growth, a moderate 20\% YoY improvement in inference efficiency), that overhead remains under 3\% at the end of 5 years.
+See Figure~\ref{fig:llm:growth}.
+
+These results are consistent with other analyses that attempt to put LLM use in the context of other computer activity~\cite{epoch2025howmuchenergydoeschatgptuse,ritchie:chatgpt}.
+
 \subsection{Sensitivity analysis}
 \label{sec:sensitivity:analysis}
+
+The results above are robust under remarkably different assumptions.
+
+For example, if we (1) massively reduce the set of end-user devices in scope to just those owned and managed by the university (about 10,000 laptops and monitors, 2,400 mobile phones), (2) triple the LLM per-query energy use to 3 Wh, and (3) assume that each of the 38,000 people at UZH uses a GPT-4o-class LLM 40 times per day, then emissions from AI still remain around 10\% of total (Figure~\ref{fig:sens1}).
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\textwidth]{fig-sens1.png}
+  \caption{A scenario with only university-managed end-user devices}
+  \label{fig:sens1}
+\end{figure}
+
+Similarly, if we cut our estimate for the embodied emissions of every type of device to the lowest value found in the literature (as opposed to a more representative middle-of-the-road value), but keep all other parameters at their default values, embodied emissions {\em still} dominate and LLM emissions continue to be a small fraction of the total, as seen in Figure~\ref{fig:sens2}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\textwidth]{fig-sens2.png}
+  \caption{A scenario with minimal embodied emissions}
+  \label{fig:sens2}
+\end{figure}
+
+Only if we combine the ``low embodied emissions'' scenario with the ``pessimal LLM'' scenario does generative AI become a major factor, contributing 18\% of total emissions.
 

### 2025-08-04 doc/paper.tex
Recommendations

@@
 
-First, do not panic about LLMs. Their emissions comprise a small fraction of total IT emissions, they will get more efficient, and attempting to curb their use is likely to be a losing battle.
+First, do not panic about the potential impact of generative AI.
+Despite the popular narratives of breathtaking data center growth, LLM emissions comprise only a small fraction of per-capita IT emissions, and are likely to continue to do so in the foreseeable future.
+While training costs are likely to continue to grow, they will also likely be amortized over longer useful lifetimes, since the models will be more capable and AI companies will need to recoup their financial investments.
+Additionally, inference is likely to continue to become more efficient as hardware and software both improve.
 
-Invest in infrastructure for the long term.
-Run awareness campaigns and adapt IT policies so that personal devices and university IT systems are used for as long as possible.
-Amortize manufacturing emissions over the longest period possible.
+Second, management of physical IT assets (workstations, monitors, etc.) is probably an underappreciated and underutilized lever for emissions reduction.
+Total emissions for these devices are dominated by manufacturing and disposal, so extending their life by a certain amount leads to an almost reciprocal reduction in emissions.
+For example, if we start the model with default parameters and then increase laptop lifetime from 4 to 6 years and mobile phone lifetime from 3 to 4 years, total campus emissions decrease from 3.9 M\tcoe\ to 3.2 M\tcoe, an 18\% reduction!
 
-Improve the cooling efficiency of university data centers, or engage cloud vendors and academic compute centers (such as CSCS) to validate the potential benefits of a migration to cloud.
+Note that an effective, large-scale extension of IT device lifetime is probably not just a matter of IT policy but of communication and education.
+The university could reach out to students and staff to sensitize them to these issues, provide more support for device maintenance and repair, etc.
+
+Third, migration to a commercial cloud may help reduce emissions by improving cooling efficiency relative to on-campus data centers.
+However, the reduction in cooling emissions in the best case is probably moderate ($\le 5$\% of total IT emissions), and it should be weighed against the emissions of cloud servers that would replace on-premise servers.
+An exact accounting requires detailed work with the cloud vendor: it is of course possible that the servers themselves would be more efficient, further increasing the margin of improvement.
+However, decision makers should also consider whether the costs of a cloud migration could be redirected to more cost-effective emission reduction actions (possibly unrelated to IT).
 
@@
 \section{Conclusion}
+
+We have presented a simple but powerful model
 

### 2025-08-05 doc/paper.tex
Related work

@@
 Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}.
-At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 M\tcoe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
+At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 k\tcoe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}.
 The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).
@@
 
-The energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,epoch2024optimallyallocatingcomputebetweeninferenceandtraining,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring,epoch2025howmuchenergydoeschatgptuse}.
-
 A widely cited 2023 study~\cite{devries2023growing} estimated the average energy consumption of a large language model like ChatGPT at approximately 3 Wh / request, about ten times the energy of a standard Google search query.
@@
 Total emissions for these devices are dominated by manufacturing and disposal, so extending their life by a certain amount leads to an almost reciprocal reduction in emissions.
-For example, if we start the model with default parameters and then increase laptop lifetime from 4 to 6 years and mobile phone lifetime from 3 to 4 years, total campus emissions decrease from 3.9 M\tcoe\ to 3.2 M\tcoe, an 18\% reduction!
+For example, if we start the model with default parameters and then increase laptop lifetime from 4 to 6 years and mobile phone lifetime from 3 to 4 years, total campus emissions decrease from 3.9 k\tcoe\ to 3.2 k\tcoe, an 18\% reduction!
 
@@
 
+Emissions modeling of information technology at organizational scales remains underdeveloped.
+Universities seem to lack standardized frameworks for measuring IT-related carbon footprints, and our work is in part an effort to help close this gap.
+
+Many carbon footprint studies treat IT as a minor subcomponent, despite IT infrastructure representing 5-10\% of total institutional emissions.
+For example, \textcite{herth:quantifying:2022} make a comprehensive carbon assessment of TU Delft and find that ICT constitutes 8\% of emissions, but do not break that down further, making it difficult to know how to improve.
+\textcite{helmers:universities:2021} exclude IT impacts from their 20-university comparison study, claiming that ``almost all universities worldwide will be working with quite similar IT technology''.
+\textcite{kiehle:2023} include IT equipment in a carbon accounting of the University of Oulu but limit themselves to lifecycle analysis of equipment purchased in the last 3 years.
+
+We are not alone in struggling with data availability problems, including incomplete asset inventories, sparse energy consumption data, etc.
+\textcite{jurgens:2023} writes that ``data availability is the main limiting factor for a holistic assessment'' of higher education institutions (not limited to IT).
+
+Efforts to standardize IT emissions measurement and reporting are advancing, \eg, through standards such as ITU-T L.1420, but they remain fragmented~\cite{wb:itu:ict}.
+Our own research on embodied and operational emissions revealed massive variability in carbon reporting among manufacturers and across studies: system boundaries range from cradle-to-gate through cradle-to-grave, functional units vary from per-device to per-computation, and values for the same parameter might vary by a factor of 2 or more.
+
+In this study, our analysis has been bottom up, building a model of the university's IT emissions from individual estimates of devices, queries, etc.
+There are important tradeoffs between bottom-up precision and top-down comprehensiveness~\cite{hilty:2009:lca}.
+Bottom-up approaches provide higher accuracy for specific sources and transparency through clear traceability.
+They also make it easier to identify specific improvement areas.
+However, they require extensive primary data collection, require up-front decisions about what components to include or not, and may miss system-wide effects.
+Conversely, top-down approaches offer efficiency and a broader system perspective, but sacrifice the precision and granularity necessary for operational decision-making.
+
+The field of AI is evolving rapidly and the main AI vendors are secretive about the architecture and scale of their models.
+This has led to many efforts to estimate the footprint of LLMs, from academic papers to blog posts~\cite{devries2023growing,heguerte2023estimate,luccioni2022estimating,ritchie:chatgpt,epoch2025howmuchenergydoeschatgptuse}.
+More generally, the energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,epoch2024optimallyallocatingcomputebetweeninferenceandtraining,gowda2024watt,harding2024watts,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring}.
+\textcite{lacoste:mlco2} and \textcite{faiz2024llmcarbon} have developed more systematic approaches to estimate the footprint of models based on factors like electricity emissions intensity, hardware specifications, and network architecture.
+
+In a provocative and maybe dystopian paper, \textcite{tomlinson:2024:carbon} argue that AI has lower emissions than humans for various writing and illustrating tasks.
+Be that as it may, it would fall into the category of indirect effects of AI, which we do not attempt to cover in this paper.
+
 \section{Conclusion}
@@
 We have presented a simple but powerful model
+
+The proposed research addresses multiple critical gaps: developing consistent methodological frameworks for institutional IT and AI assessment, creating granular modeling approaches that capture technological heterogeneity, integrating embodied and operational emissions across full technology lifecycles, and establishing standardized metrics for institutional comparison and benchmarking.
+By synthesizing insights from LCA methodologies, AI emissions research, and institutional sustainability practices, this work can contribute substantially to both academic understanding and practical institutional climate action capabilities.
 

### 2025-08-05 doc/paper.tex
Conclusion

@@
 
-We have presented a simple but powerful model
+We have presented a simple but powerful model to estimate the carbon footprint of IT infrastructure and IT-related activities at the University of Zürich.
+While the concrete parameter values we chose are specific to UZH, we have attempted to create a framework that can be applied to other universities (and maybe other large campus facilities) and can help provide a basis for evaluation and planning.
 
-The proposed research addresses multiple critical gaps: developing consistent methodological frameworks for institutional IT and AI assessment, creating granular modeling approaches that capture technological heterogeneity, integrating embodied and operational emissions across full technology lifecycles, and establishing standardized metrics for institutional comparison and benchmarking.
-By synthesizing insights from LCA methodologies, AI emissions research, and institutional sustainability practices, this work can contribute substantially to both academic understanding and practical institutional climate action capabilities.
+Our model has broad system boundaries: it includes both university- and student-owned devices, cloud and supercomputing facilities, and generative AI services.
+Whenever possible it accounts for both embodied and operational emissions.
+The model is bottom-up and exposes almost 100 configurable parameters to support different environments and enable exploration of different scenarios.
+It is available at \url{https://maxp.net/it-emissions} and can be freely modified under an MIT License.
 
-Our analysis shows that manufacturing (“embodied”) emissions dominate UZH's IT carbon footprint, while operational impacts, including AI adoption, are relatively modest.
+Within the context of UZH, our analysis shows that embodied emissions dominate the IT carbon footprint, while operational impacts, including large-scale adoption of generative AI, are relatively modest.
+In particular:
 
@@
 
-  \item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint.
-For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ \tcoe\ annually.
-Usage-related emissions amount to about a tenth of that.
+  \item Under the assumption that a large majority (80\%) of UZH students and staff have a laptop and smartphone, the manufacture of personal devices represents about $\frac{2}{3}$ of the university IT carbon footprint.
+Operational emissions account for only 15--20\% of total emissions of end-user devices (laptops, monitors, phones, etc.).
 
-  \item Widespread adoption of LLMs and other frontier models should have moderate impact on total emissions.
-If everyone at UZH made heavy use of AI (15 queries daily), this would add approximately 220 t CO2e annually, of which 160 t directly attributable to queries.
-While this would almost double an individual's real-time energy use, the impact on overall emissions is relatively small.
+  \item Embodied emissions of end-user devices account for almost half of total emissoins even we consider only university-owned devices (equivalent to about $\frac{1}{4}$ of the population).
 
-  \item Embodied emissions are an important consideration for university-owned devices as well.
-If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ \tcoe, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.
+  \item Data center cooling and efficiency represent a potential savings opportunity.
+Inefficient cooling in on-premise data centers causes for 150--200 t \coe\ / year of emissions that could be avoided by a migration to a commercial cloud facility.
+However, any such move requires in-depth analysis with the cloud vendor to ensure that it is still a net savings after accounting for the different server infrastructure and other changes.
 
-  \item Data center cooling and efficiency represents a potential savings opportunity.
-The largest operational emissions are from data center servers and cooling (320 t).
-UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders.
-Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ \tcoe\ annually, though this estimate requires further validation.
+  \item Widespread adoption of LLMs and other frontier models should have only moderate impact on total emissions.
+Today LLM use probably causes about 1\% of total IT emissions, even when accounting for training overhead and the LLM's embodied emissins.
+Even assuming very aggressive growth, generative AI is unlikely to exceed 5--10\% of overall IT emissions five years from now.
 
 \end{itemize}
+
+These results appear to be robust even when shifting parameter values to the extremes of their likely ranges (\eg, low embodied emissions, high per-query AI overhead, etc.) and lead us to make some high-level IT policy recommendations in Section~\ref{sec:recommendations}.
 

### 2025-09-09 doc/paper.tex
Nit
Increase university-owned laptop/monitor lifetimes per new central IT data, and adjust recommendations/conclusions.

@@
 \newcommand{\coe}{CO$_2$e}
-\newcommand{\gcoe}{g \coe}
-\newcommand{\tcoe}{t \coe}
+\newcommand{\gcoe}{g~\coe}
+\newcommand{\tcoe}{t~\coe}
 \newcommand{\kgcoe}{k\gcoe}
@@
 \title{Modeling the impact of IT- and AI-related emissions in a university setting}
-\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}
+\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch}}
 
@@
     \textbf{Device type} & \textbf{Count} & \textbf{Life (y)}\\ \hline
-    Laptops & 10,440 & 4 \\
-    Monitors & 10,400 & 6 \\
+    Laptops & 10,440 & 8 \\
+    Monitors & 10,400 & 8 \\
     Servers & 1,121 & 5 \\
-    Printers & 333 & 6 \\
+    Printers & 333 & 8 \\
     Mobile phones & 2,439 & 3 \\
-    A/V devices & 431 & 6 \\ \hline
+    A/V devices & 431 & 8 \\ \hline
   \end{tabular}
@@
 Traditional desktop computers appear to be infrequent on campus. By default we set their number to zero in our model, but we still provide the framework to model them if desired.
+
+The device lifetime for university-managed laptops and monitors comes from UZH's central IT department and is remarkably long.
+For other types of devices we assume a similarly conservative management of hardware resources.
+We also estimate that student-owned laptops have a shorter lifetime of 5 years, and therefore, for the purposes of the results later in this paper, we set the default average lifetime of laptops in our model to 6 years.
 
@@
 
-Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays, (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day.
+Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays (250 days per year), (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day.
 Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.
@@
 
-Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 \tcoe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
+Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions and mean device lifetimes to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 \tcoe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different.
 Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.
@@
 
-Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170 \tcoe\ of cooling overhead, and replaces 415 \tcoe\ of total server emissions with 380 \tcoe\ of cloud emissions.
+Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170~\tcoe\ of cooling overhead, and replaces 415~\tcoe\ of total server emissions with 380~\tcoe\ of cloud emissions.
 See Figure~\ref{fig:cloud}.
@@
 
-For example, if we (1) massively reduce the set of end-user devices in scope to just those owned and managed by the university (about 10,000 laptops and monitors, 2,400 mobile phones), (2) triple the LLM per-query energy use to 3 Wh, and (3) assume that each of the 38,000 people at UZH uses a GPT-4o-class LLM 40 times per day, then emissions from AI still remain around 10\% of total (Figure~\ref{fig:sens1}).
+For example, if we (1) massively reduce the set of end-user devices in scope to just those owned and managed by the university (about 10,000 laptops and monitors, 2,400 mobile phones), (2) triple the LLM per-query energy use to 3 Wh, and (3) assume that each of the 38,000 people at UZH uses a GPT-4o-class LLM 40 times per day, then emissions from AI are still just 11\% of total (Figure~\ref{fig:sens1}).
 
@@
 
-Only if we combine the ``low embodied emissions'' scenario with the ``pessimal LLM'' scenario does generative AI become a major factor, contributing 18\% of total emissions.
+Only if we combine the ``low embodied emissions'' scenario with the ``pessimal LLM'' scenario does generative AI become a substantial factor, contributing 16\% of total emissions.
 
@@
 Total emissions for these devices are dominated by manufacturing and disposal, so extending their life by a certain amount leads to an almost reciprocal reduction in emissions.
-For example, if we start the model with default parameters and then increase laptop lifetime from 4 to 6 years and mobile phone lifetime from 3 to 4 years, total campus emissions decrease from 3.9 k\tcoe\ to 3.2 k\tcoe, an 18\% reduction!
+For example, if we start the model with default parameters and an industry-standard device lifetime of 4 years for laptops and 3 for mobile phones, then extend those to 6 and 4 years, respectively, total campus emissions decrease from 3.9 k\tcoe\ to 3.2 k\tcoe, an 18\% reduction!
+To its credit, the University of Zürich central IT department already keeps devices in service for about 8 years on average, so UZH may not have much room for improvement in this regard.
 
 Note that an effective, large-scale extension of IT device lifetime is probably not just a matter of IT policy but of communication and education.
-The university could reach out to students and staff to sensitize them to these issues, provide more support for device maintenance and repair, etc.
+Universities could reach out to students and staff to sensitize them to these issues, provide more support for device maintenance and repair, etc.
 
@@
 
-  \item Under the assumption that a large majority (80\%) of UZH students and staff have a laptop and smartphone, the manufacture of personal devices represents about $\frac{2}{3}$ of the university IT carbon footprint.
+  \item Under the assumption that a large majority (80\%) of UZH students and staff have a laptop and smartphone, the manufacture of personal devices represents almost $\frac{2}{3}$ of the university IT carbon footprint.
 Operational emissions account for only 15--20\% of total emissions of end-user devices (laptops, monitors, phones, etc.).
 
-  \item Embodied emissions of end-user devices account for almost half of total emissoins even we consider only university-owned devices (equivalent to about $\frac{1}{4}$ of the population).
+  \item Embodied emissions of end-user devices account for almost half of total emissions even we consider only university-owned devices (equivalent to about $\frac{1}{4}$ of the population).
 

### 2025-09-11 doc/paper.tex
Change citation style for journal
Add brief discussions of Larsen and Garcia papers.

@@
-% boavizta mention immaturity of assessments and variability of estimates
-
 \documentclass[11pt]{article}
@@
 \usepackage{libertine}
+\usepackage{graphicx}
+\usepackage{hyperref}
+\hypersetup{
+    hidelinks
+}
+\usepackage{titlesec}
+\usepackage{enumitem}
+\usepackage{titling}
+
+\usepackage[style=authoryear]{biblatex}
+\addbibresource{ref.bib}
+\let\cite\parencite
+
 \renewcommand{\sfdefault}{fvs} % Use as sans-serif
@@
 \newcommand{\gcoekwh}{\gcoe\ / kWh}
-%\usepackage{helvet} % Helvetica font for sans-serif
-%\usepackage{courier} % Courier font for monospace
-\usepackage{graphicx}
-\usepackage{hyperref}
-\hypersetup{
-    hidelinks
-}
-\usepackage[style=numeric]{biblatex}
-\usepackage{titlesec}
-\usepackage{enumitem}
-\usepackage{titling}
 
@@
 
-\addbibresource{ref.bib}
 \title{Modeling the impact of IT- and AI-related emissions in a university setting}
-\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch}}
+\author{Massimiliano Poletto\thanks{Independent researcher, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch}}
 
@@
 
+Studies such as \textcite{larsen:2013} assess the carbon footprint of a university but do not consider IT-related emissions specifically.
+For example, Larsen et al.
+lump IT with ``equipment'', including machinery and science and technology equipment.
+
 Many carbon footprint studies treat IT as a minor subcomponent, despite IT infrastructure representing 5-10\% of total institutional emissions.
@@
 In this study, our analysis has been bottom up, building a model of the university's IT emissions from individual estimates of devices, queries, etc.
+Some other studies, including \textcite{larsen:2013} and \textcite{garciaalaminos:2022}, have used top-down approaches based on environmental input-output (EIO) models.
 There are important tradeoffs between bottom-up precision and top-down comprehensiveness~\cite{hilty:2009:lca}.
@@
 They also make it easier to identify specific improvement areas.
-However, they require extensive primary data collection, require up-front decisions about what components to include or not, and may miss system-wide effects.
+However, they require extensive primary data collection, force up-front decisions about what components to include or not, and may miss system-wide effects.
 Conversely, top-down approaches offer efficiency and a broader system perspective, but sacrifice the precision and granularity necessary for operational decision-making.
